### 2019-06-03 paper.tex
add paper

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
 \geometry{
 top=30mm,
 left=1.5in,
 right=1.5in
 }
\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}

\title{Kratos: blockchain solution for data privacy, literacy, and student agency in a data-driven educational ecosystem}
\author{Dr.\ Velislava Hillman \quad Varunram Ganesh \quad Geoffrey Martin}
\date{May 2019}

\begin{document}

\maketitle

\textit{Abstract}
Growing digitization has made data ownership an important focus point for institutions and students. Broadly, there are three issues, which require urgent attention for optimization of data privacy, literacy, and utilization. First, schools globally are equivocal about data generated by and about students as a result of the digitization of instruction, learning, and assessment. They lack necessary frameworks for data literacy, data interoperability, and optimization while maintaining privacy and control. Second, the scale, source, and nature of school data makes its interoperability impractical, resulting in an inability to assess the true impact of educational technologies on instruction and learning. Third, while data helps teachers improve pedagogical practices, an increasingly data-driven decision-making process suggests that student dimensions of learning and equitable participation in curriculum design becomes secondary. Finding a balance between increasingly data-driven decision making and student voice and choice is critical for an efficient educational ecosystem. In this paper, we introduce Kratos: an immutable decentralised data management system that provides data privacy and applied data literacy while empowering students with a user interface for data governance and active participation in the educational ecosystem. Using the advantages afforded by blockchain technologies, Kratos provides easy authentication, confidentiality, and accountability for data access, use, and sharing. The objective of Kratos is thus to equip schools and students with the ability to access, manage, control, and understand data and how, why, and by whom data is accessed and used at any one point in time without compromising student agency and privacy. This paper describes proof of concept for a decentralised data management system and its benefits to the educational ecosystem.
\bigbreak

\section{Introduction}

To improve work, school practitioners at local, district, and federal levels need data interoperability and educational information from data mining [Gates Foundation, 2015; US Dept of Education, 2012; ACT Policy Report, 2015]. This need originally emanates from the United States federal law which mandates schools at district, state, and federal level to collect student data to serve as 'accountability' metric for assessing school effectiveness [Elementary and Secondary Education Act, 2001; 2010].
\bigbreak
Educational data is important not only for evaluating school quality. It can also contribute to research in learning and advance theory [Baker, 2016]. It can help guide intervention and lead to higher school attainment [Arnold and Pistill, 2012], improved pedagogy [Baker, 2014], and better parental involvement in school matters [Hawn, 2015]. A common goal of educational data is to equip educators with key information that they can act upon and use to the benefit of learners [Baker, 2014]. Educational data can enable an applied data literacy. Furthermore, access to school records provides an opportunity for student agency by enabling learning and participation.
\bigbreak
On the other hand, concerns arise about the pervasiveness of fine-grained data collected about students [Zeide, 2017] and the risk of creating a 'permanent record' that can impede upon learners' futures [Cody, 2013]. The use of online platforms, programs, and applications in schools often provided by for-profit vendors, many of whom have unclear policies about data privacy and third-party sharing [Common Sense Education, 2018], generates continuous stream of data about students' behavior and performance at an unprecedented scale [Zeide, 2017]. Big data complicates traditional understanding of what constitutes sensitive information and what information serves an educational purpose. Data continues to drive decision-making of practitioners, vendors, and policy makers, marginalizing student dimensions of learning and equitable participation in the curriculum design.
\bigbreak
Interoperability challenges between vendors and schools at local, district, or federal level, further pose barriers not only to cohesive data management and sharing but also lacks technological privacy infrastructure and accountability [Zeide, 2014; Common Sense Media Research, 2018]. While digitizing and collecting student data is not new [Fitzgerald, 2014], technological developments like cloud computing and the growing use of web applications, learning management systems that gather fine-grained student data, amplify the concerns about data transfer, storage, use, and analysis [Sultan, 2010].
\bigbreak
Increasingly digitized learning environments enable constant data collection and algorithm-based assessment which change core school functions of teaching, assessment, and credentialing [Zeide, 2017]. Digital systems for teaching and assessment drive data-based decision-making and enable micromanagement of students that can further pose restrictions on teacher autonomy and student and parent participation and willingness to challenge education decision-making. Interactive learning environments allow students to manage coursework online, take tests, and use a variety of online learning material. Schools deploy management systems for sorting and storing reporting and operational data. While some data is stored on local servers at school level, other school data is managed by third-party vendors [Zhang, Li, and Hao, 2015], who may consider data proprietary and thus inaccessible. This puts additional barriers to effective data sharing and comprehensive understanding of school data.
\bigbreak
The present challenges of fragmented data access and use, lack of student agency, accountability and auditability to data access and use demand a socio-technological solution, one which can ensure that educational data can be used to help improve school processes while student privacy, agency, and future opportunities are not diminished.
\bigbreak
In this work we propose the development of a blockchain structure for student data management, privacy, accountability, and auditability. We optimise on available data standards and construct references to disparate data across different systems and applications on a decentralised ledger. Our system organises these references onto an integral structure of student data and complements the existing educational data standards and interoperability in three distinctive ways. Through on-the-network permissioning, we enable school and student agency with data accuracy, auditability, and sharing. We design modular APIs to integrate with existing school systems and data standards. Lastly, we build a simple user interface that gives students/parents and schools access control of records enabling control over what portion or part of the data can be shared. Additionally, the interface enables applied data literacy and an application with social functionality for student feedback.
\bigbreak
We present Kratos not as the only solution to lack of data interoperability, accountability, and privacy but rather as a system that complements existing efforts and solutions developed by the various stakeholders in the educational ecosystem (Data Quality Standards; Project Unicorn).

\section{The need for a solution}

\subsection{Student agency and participation}
The UNESCO framework for educational planning states that "the concern of planners is twofold: to reach a better understanding of the validity of education in its own empirically observed specific dimensions and to help in defining appropriate strategies for change" (Haddad, 1995, pp. 5-6).

While summative and cumulative assessments provide "empirically observed specific dimensions" about student academic performance, student agency and active participation is equally required in order for policy and education to design "strategies for change". Growing use of education technologies enables data-driven decision-making [Gibson et al., 2015]. Technology-mediated instruction and assessment tools with learning analytics functionality track and diagnose student progress. Most educational technologies can interpret and equip educators with information via digital dashboards and 'skill meters'[Baker, 2016; New, 2016]. Fine-grained and continuously accumulated data about student behavior and performance surpasses traditional notions of assessment [Zeide, 2017] posing limitations over student dimensions of learning and equitable participation in the curriculum design.

Our prototype provides a graphical interface for student involvement in and accessibility to school data. While students and equally their parents can become acquainted with any changes and meanings of school data, students can also participate with personal feedback to the learning process. Student participation with personal perspectives and reflection are integral to the learning process. Such reflection is invariably carried out through questionnaire surveys examining school climate [Holahan and Batey, 2019]. Both school climate surveys and the proposed student feedback application provide flexible selection of questionnaires and measurements with a common goal to improve school climate. However, the proposed application enables not only feedback from older students [Ibid.] but from all students, provided that the feedback addresses the common goal. The application further enables student agency and control over the frequency of feedback.

\subsection{Data literacy}
Data literacy is the ability to understand, create, and communicate data as information. Beyond that, in an increasingly data-driven socio-economic systems it becomes imperative not only to understand data but interact with it and participate in the decision-making processes it increasingly exerts its influence. While data provides insight about how systems operate, the inferences drawn from it is without context. Fine-grained data collected over long periods of time can lead to negative impact on individual well-being and pose limitations over future opportunities [Gasser et al., 2018]. The growing digitization of schools create learning environments that enable constant data generation, the access to and use of which becomes hard to control, audit, and account for. The risk of digitized school environments constantly generating data and algorithmic assessments becoming permanent record that at any one point in time may limit individual opportunities leads us to prioritise on developing a socio-technological solution that enables applied data literacy for students at a young age. \textit{a note on Parents for Privacy Consortium and Data Quality Standards who make such efforts to educate students and parents about school data and that we (Kratos) complement these efforts.}

\section{Existing problems}
In order to have a better picture on the problems schools currently face, Kratos has partnered with the Cambridge Public School (CPS) District, the controlling body for all schools in Cambridge, Massachusetts. With the help of the Chief Information Officer, Chief Technical Officer and Database Administrator, we explored problems and ways to solve them along with feedback on potential mitigation schemes:

\begin{itemize}
  \item Data access - CPS has agreements with over 150 partners ([7]) and on account, none of these providers provide direct access to data. Data access is instead provided in the form of a report or a summary which the district database administrator can request and download. Some providers are even less flexible and only offer to give reports once in 45 days making it impossible for teachers to act on feedback that they might infer from the data.
  \item Lack of standards - The different vendors that CPS partners have their own implementation, most of them not regualted by any standards. The schools themselves follow a common standard defined by the SIF but there is no way to enforce this upon vendors. Even if the vendor agreed to follow a standard defined by the SIF, it is not possible to concretely verify this claim.
  \item Lack of transparency - The vendors do not provide a comprehensive list of the data they collect from students. This makes data audits extremely difficult since we don't exhaustively know what to audit.
  \item Security - It is impossible to verify if vendors are following best practices in data storage and retrieval. This is extremely dangerous since a malicious entity can arbitrarily change or retrieve student records from a poorly maintained database
\end{itemize}

We can thus see that schools are finding it increasingly difficult to obtain data about their own students and are becoming isolated due to a lack of knowledge on how vendors are processing this data.

\section{Kratos}
Kratos aims to define a set of principles and guidelines that schools should follow and provides a platform for both schools and students to see what kind of data is being collected by vendors along with ways to tweak data access. The goal for students would be to see how their data is being used and above the age of 18, gain acccess to this data. We define this as Data Ownership and Visibility- how students can own the data after the age of 18 and how students can see what their data is being used for. The goal for the school would be to make disparate data from different vendors compatible with each other and we define this problem as Data Interoperability.

\subsection{Data ownership}
In conventional systems, data ownership can be proved with the help of an access token but this provides no guarantee on when the owner came into possession of the data. In order to attest ownership at a specific point in time, we need timestamping services like those described in \hyperref[sec:1]{[1]}, \hyperref[sec:2]{[2]} and \hyperref[sec:3]{[3]}. A timestamping service requires something to be committed along with the timestamp and Kratos envisions this to be a cryptographic hash which also acts as an access token which vendors can use to access the data.
\bigbreak
The aim of using a cryptographic hash like SHA-3 \hyperref[sec:6]{[6]} is to ensure a uniquely random reference to the data to avoid out of channel data leaks. Kratos suggests using an element of randomness like a salt to generate the hash in order to have the ability to revoke tokens by regenerating randomness. Kratos enforces that all schools encrypt their data before creating access tokens to mitigate the risk of loss/theft of data. Past studies like \hyperref[sec:4]{[4]} and \hyperref[sec:5]{[5]} show that firms are willing to circumnavigate laws to collect data and encrypting data by design ensures that no third party can have access without being granted so explicitly.
\bigbreak
Kratos does not specify at which level data needs to be encrypted - it can be at the school level or at the student level depending on existing frameworks and rules surrounding the school. If the school chooses to encrypt student data on behalf of the student, Kratos enforces that the school use a unique key for each individual to minimize the risk of key theft. In addition to this, Kratos suggests that schools store their encrypted data in a distributed file storage system like IPFS to ensure data redundancy and availability in case of a mishap. Storing data on IPFS also makes it easier to create timestamps since it is sufficient to reference the IPFS pointer instead of potentially hashing the whole data.
\bigbreak
In the event a user wants to revoke access to a particular vendor, he could do so by changing the encryption key or by changing the randomness used to generate the access token. We suggest users don't regenerate encryption keys but Kratos does not enforce this and will provide end users with the option to choose between the two.
\bigbreak
Since schools have their own sets of policies, Kratos doesn't strictly enforce a set of practices for users to follow. This ensures that adoption of Kratos isn't constrained by a certain set of rules. At the same time, Kratos defines a set of minimum requirements to be on board to ensure good practices on date protection are followed.
\bigbreak
Kratos also doesn't specify how and where the commitments need to be stored and leaves it to schools to provide their feedback. Potential solutions include a centralized timestamping server, a permissioned blockchain with the different schools as the nodes and simple timestamping commitments to an existing blockchain. All three options have their benefits and constraints and Kratos would arrive at a final recommendation depending on the option the majority of schools are convenient with
\bigbreak

\subsection{Data Interoperability}
A traditional approach to interoperability would be to force a given set of standards on third party vendors, so that they reference them. However, work done by standardization bodies like SIF and CEDS has shown that this is at best partially effective since there is no financial or social incentive for vendors to migrate to the proposed set of standards. Kratos attempts to solve this issue by proposing a solution where different fields described by different vendors can be mapped to a single underlying scheme defined by Kratos and third party vendors would not have to worry about using disparate fields.

\bigbreak
As an example, let us assume the underlying fields defined by Kratos are Grade\_A, Grade\_B and Grade\_C. If a specific vendor has a different name for the same field (example GradeA, GradeB and GradeC), Kratos would ingest the data from the vendor defined fields and convert it to the ones defined by Kratos. This should be done across different formats since the reports database administrators receive right now are formatted as JSON, CSV or are in the form of Excel sheets.

\bigbreak
Kratos proposes that this be done in the form of a templated script that can be written for every vendor. Building a common script for all the vendors poses a challenge since the number of vendors and the different fields and standards they follow are ever changing. This templated script would be trigerred automatically by Kratos each time it receives an incoming report and the report itself will be parsed ot udnerstand which vendor it originates from.

\bigbreak
The fields that Kratos defines will be adapted from the existing SIF standard followed by schools in Cambridge and would also have routine input from various experts on the subject. The number of fields however would change since SIF has over 700 different fields and it is not possible to accurately map all these fields. Kratos defines the idea of a "bucket" - a collection of data fields collated together as a single entity to make it easier for students and administrators to monitor them. Common examples of buckets would be PII (Personally Identifiable Information), Grade Reports (containing grade reports for each subject) and Attendance Records (recording the class wise attendance of the student). More such buckets will be added after receiving input from students and teachers on what the best ways to classify this data are.

\bigbreak
The models and scripts developed as part of enabling interoperability will be open source and subject to continuous updates. We would also be having a complete code audit before relasing the system in a production environment to ensure the model performs and behaves the way it was designed and intended to.

\bigbreak
The buckets defined by Kratos can be used in multiple ways: students can have a better understanding of how their data is being used, parents can know hwat kind of data is being collected by their children and administrators can ensure that sensitive information is not being shared with third party vendors. After the age of 18, this control would be given to the hands of students and they can decide if they wish to continue sharing data with these parties.

\subsection{User interface}
Visuals, wireframes, mockups, description of these

\section{Discussion and future work}
In this paper we propose Kratos, a data management system, which enables data interoperability for otherwise disparate data, applied data literacy, student agency over their data and participation in designing their education.

- a few paragraphs on data and learning analytics..

- paragraph on Kratos as an add-on, rather than replacement of existing efforts:

\section{References}
\begin{enumerate}
    \label{sec:1}
    \item Todd, P. OpenTimestamps: Scalable, Trustless, Distributed Timestamping with Bitcoin (2016). URL \url{https://petertodd.org/2016/opentimestamps-announcement}
    \label{sec:2}
    \item Szalachowski, P. (2018). Towards more reliable Bitcoin timestamps. arXiv preprint arXiv:1803.09028.
    \label{sec:3}
    \item Massias, H., Avila, X. S., \& Quisquater, J. J. (1999). Design of a secure timestamping service with minimal trust requirement. In the 20th Symposium on Information Theory in the Benelux.
    \label{sec:4}
    \item Zetter, K. Google Collected Data on Schoolchildren without permission (2016). URL \url{https://www.wired.com/2015/12/google-collected-data-on-schoolchildren-without-permission/}
    \label{sec:5}
    \item Dissent, Back-To-School Revolt in Springfield? Employees balk over using Google Drive as evidence of massive privacy breach mounts (2018).  URL \url{https://www.pogowasright.org/back-to-school-revolt-in-springfield-employees-balk-over-using-google-drive-as-evidence-of-massive-privacy-breach-mounts/}
    \label{sec:6}
    \item Bertoni, G., Daemen, J., Peeters, M., Van Assche, G., \& Van Keer, R. (2012). Keccak implementation overview. URL  \url{http://keccak.neokeon.org/Keccak-implementation-3.2.pdf}.
    \label{sec:7}
    \item CPSD. District Agreements Listing (2019). URL  \url{https://sdpc.a4l.org/district_listing.php?districtID=457}
    \label{sec:8}
    \item Colins, A., and Halverson, R. (2018).\textit{ Re-thinking education in the age of technology: The digital revolution and schooling in America}. Teach Zeide, E. (2016).
    \label{sec:9}
    \item 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?.
    \label{sec:10}
    \item Colins, A., and Halverson, R. (2018).\textit{ Re-thinking education in the age of technology: The digital revolution and schooling in America}. Teachers College Press: New York.
    \label{sec:11}
    \item Cody, A. (2013). Will the data warehouse become every student and teacher's 'permanent record'? \textit{Education Week}, May, 20.
    \label{sec:12}
    \item Gibson, D. C., Webb, M., and Ifenthaler, D. (2015). Challenges of big data in educational assessment.\textit{Proceedings of the IADIS International Conference of Exploratory Learning in Digital Age}, 92-100.
    \label{sec:13}
    \item Sultan, N. (2010). Cloud computing for education: a new dawn? 30 International Journal of Information Management  109.
    \label{sec:14}
    \item Fitzgerald, B. (2014). Data collection isn't new. And it predates common core.Funny Monkey.
    \label{sec:15}
    \item Zeide, E. (2016). 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?.
\end{enumerate}
\end{document}

### 2019-06-03 paper.tex
add images and changes

@@
 \usepackage{geometry}
+\usepackage[final]{graphicx}
+\usepackage{grffile}
  \geometry{
@@
 
-To improve work, school practitioners at local, district, and federal levels need data interoperability and educational information from data mining [Gates Foundation, 2015; US Dept of Education, 2012; ACT Policy Report, 2015].
+To improve work, school practitioners at local, district, state, and federal levels need data interoperability and educational information from data mining [Gates Foundation, 2015; US Dept of Education, 2012; ACT Policy Report, 2015].
 This need originally emanates from the United States federal law which mandates schools at district, state, and federal level to collect student data to serve as 'accountability' metric for assessing school effectiveness [Elementary and Secondary Education Act, 2001; 2010].
@@
 \bigbreak
-Interoperability challenges between vendors and schools at local, district, or federal level, further pose barriers not only to cohesive data management and sharing but also lacks technological privacy infrastructure and accountability [Zeide, 2014; Common Sense Media Research, 2018].
+Interoperability challenges between vendors and schools at local, district, state, or federal level further pose barriers not only to cohesive data management and sharing but also lacks technological privacy infrastructure and accountability [Zeide, 2014; Common Sense Media Research, 2018].
 While digitizing and collecting student data is not new [Fitzgerald, 2014], technological developments like cloud computing and the growing use of web applications, learning management systems that gather fine-grained student data, amplify the concerns about data transfer, storage, use, and analysis [Sultan, 2010].
 \bigbreak
-Increasingly digitized learning environments enable constant data collection and algorithm-based assessment which change core school functions of teaching, assessment, and credentialing [Zeide, 2017].
-Digital systems for teaching and assessment drive data-based decision-making and enable micromanagement of students that can further pose restrictions on teacher autonomy and student and parent participation and willingness to challenge education decision-making.
+Increasingly digitized learning environments enable constant data collection and algorithm-based assessment which change core school functions of teaching, assessment, and accreditation [Zeide, 2017].
+Digital systems for teaching and assessment drive data-based decision-making and enable micromanagement of students that can further pose restrictions on teacher autonomy and student (and parent) participation and willingness to challenge education decision-making.
 Interactive learning environments allow students to manage coursework online, take tests, and use a variety of online learning material.
 Schools deploy management systems for sorting and storing reporting and operational data.
-While some data is stored on local servers at school level, other school data is managed by third-party vendors [Zhang, Li, and Hao, 2015], who may consider data proprietary and thus inaccessible.
+While some data is stored on local servers at school district level, other school data is managed by third-party vendors [Zhang, Li and Hao, 2015], who may consider data proprietary and thus inaccessible.
 This puts additional barriers to effective data sharing and comprehensive understanding of school data.
 \bigbreak
-The present challenges of fragmented data access and use, lack of student agency, accountability and auditability to data access and use demand a socio-technological solution, one which can ensure that educational data can be used to help improve school processes while student privacy, agency, and future opportunities are not diminished.
+The present challenges of fragmented data access and use, lack of student agency, accountability, and auditability to data access and use demand a socio-technological solution, one which can ensure that educational data can be used to help improve school processes while student privacy, agency, and future opportunities are not diminished.
 \bigbreak
-In this work we propose the development of a blockchain structure for student data management, privacy, accountability, and auditability.
-We optimise on available data standards and construct references to disparate data across different systems and applications on a decentralised ledger.
+In this work we propose the development of a structure for student data management, privacy, accountability, and auditability.
+We build on top of existing data standards and construct a common data schema for disparate data across different systems.
 Our system organises these references onto an integral structure of student data and complements the existing educational data standards and interoperability in three distinctive ways.
-Through on-the-network permissioning, we enable school and student agency with data accuracy, auditability, and sharing.
-We design modular APIs to integrate with existing school systems and data standards.
-Lastly, we build a simple user interface that gives students/parents and schools access control of records enabling control over what portion or part of the data can be shared.
-Additionally, the interface enables applied data literacy and an application with social functionality for student feedback.
+Through network permissioning and proofs of ownership on a distributed ledger, we enable data auditability and accountability.
+We design data analytic models to integrate with existing school systems and data standards.
+Lastly, we build a simple user interface that gives students, parents and schools access to disparate data, enabling control over what portion of the data can be shared.
+Additionally, we propose an application with social functionality for student feedback.
 \bigbreak
-We present Kratos not as the only solution to lack of data interoperability, accountability, and privacy but rather as a system that complements existing efforts and solutions developed by the various stakeholders in the educational ecosystem (Data Quality Standards; Project Unicorn).
+We present Kratos not as an independent solution to lack of data interoperability, accountability, and privacy but rather as a system that complements existing efforts and solutions developed by the various stakeholders in the educational ecosystem (Data Quality Standards, 2018; Project Unicorn; Bill and Melinda Gates Foundation, 2015).
+
+\section{Existing problems}
+To contextualise the complexity related to school data - the opportunities and challenges schools face as a result of the growing digitisation of operation and academic processes - Kratos has partnered with  Cambridge Public School (CPS) district that administer public elementary and high schools in Cambridge, Massachusetts, the Access for Learning (SIF), and the Student Data Privacy Consortium.
+With the help of CPS's Information and Communication Technology Services Chief Information Officer and Database Administrator, we explored data interoperability issues with them along with feedback on potential mitigation schemes:
+
+\begin{itemize}
+  \item Data access - CPS is in agreement with over 100 vendors providing education technologies ([7]).
+The majority of these would mostly provide no direct and comprehensive access to data generated about and by students who make use of their products or services.
+Where available, data access is  provided in the form of reports - a summary of information which the district database administrator can request and download.
+Some providers are even less flexible and only offer to give reports once in 45 days making it impossible for teachers to act on feedback that they might infer from the data.
+On other occasions vendors supply data directly to teachers as well as the students in the form of digital dashboards.
+Where teachers obtain data it requires further work to convert it into meaningful information upon which the teacher can adjust and plan instruction (Data Quality Campaign, 2018).
+  \item Unified data standardization challenges - The different vendors that CPS partners have their own implementation, most of them not regulated by any standards.
+The schools themselves follow a common standard defined by the SIF but there is no way to enforce this upon vendors.
+Even if the vendor agreed to follow a standard defined by the SIF, it is not possible to concretely verify this claim.
+  \item Lack of transparency - The vendors do not provide a comprehensive list of the data they collect from students.
+This makes data audits extremely difficult since we don't exhaustively know what to audit.
+  \item Security - It is impossible to verify if vendors are following best practices in data storage and retrieval.
+This is extremely dangerous since a malicious entity can arbitrarily change or retrieve student records from a poorly maintained database
+\end{itemize}
+
+We can thus see that schools are finding it increasingly difficult to obtain data about their own students and are becoming isolated due to a lack of knowledge on how vendors are processing this data.
 
@@
 
-\section{Existing problems}
-In order to have a better picture on the problems schools currently face, Kratos has partnered with the Cambridge Public School (CPS) District, the controlling body for all schools in Cambridge, Massachusetts.
-With the help of the Chief Information Officer, Chief Technical Officer and Database Administrator, we explored problems and ways to solve them along with feedback on potential mitigation schemes:
-
-\begin{itemize}
-  \item Data access - CPS has agreements with over 150 partners ([7]) and on account, none of these providers provide direct access to data.
-Data access is instead provided in the form of a report or a summary which the district database administrator can request and download.
-Some providers are even less flexible and only offer to give reports once in 45 days making it impossible for teachers to act on feedback that they might infer from the data.
-  \item Lack of standards - The different vendors that CPS partners have their own implementation, most of them not regualted by any standards.
-The schools themselves follow a common standard defined by the SIF but there is no way to enforce this upon vendors.
-Even if the vendor agreed to follow a standard defined by the SIF, it is not possible to concretely verify this claim.
-  \item Lack of transparency - The vendors do not provide a comprehensive list of the data they collect from students.
-This makes data audits extremely difficult since we don't exhaustively know what to audit.
-  \item Security - It is impossible to verify if vendors are following best practices in data storage and retrieval.
-This is extremely dangerous since a malicious entity can arbitrarily change or retrieve student records from a poorly maintained database
-\end{itemize}
-
-We can thus see that schools are finding it increasingly difficult to obtain data about their own students and are becoming isolated due to a lack of knowledge on how vendors are processing this data.
-
 \section{Kratos}
 Kratos aims to define a set of principles and guidelines that schools should follow and provides a platform for both schools and students to see what kind of data is being collected by vendors along with ways to tweak data access.
-The goal for students would be to see how their data is being used and above the age of 18, gain acccess to this data.
+The goal for students would be to see how their data is being used and above the age of 18, gain access to this data.
 We define this as Data Ownership and Visibility- how students can own the data after the age of 18 and how students can see what their data is being used for.
@@
 In conventional systems, data ownership can be proved with the help of an access token but this provides no guarantee on when the owner came into possession of the data.
-In order to attest ownership at a specific point in time, we need timestamping services like those described in \hyperref[sec:1]{[1]}, \hyperref[sec:2]{[2]} and \hyperref[sec:3]{[3]}.
-A timestamping service requires something to be committed along with the timestamp and Kratos envisions this to be a cryptographic hash which also acts as an access token which vendors can use to access the data.
+In order to attest ownership at a specific point in time, we need time-stamping services like those described in \hyperref[sec:1]{[1]}, \hyperref[sec:2]{[2]} and \hyperref[sec:3]{[3]}.
+A time-stamping service requires something to be committed along with the timestamp and Kratos envisions this to be a cryptographic hash which also acts as an access token which vendors can use to access the data.
 \bigbreak
@@
 Kratos also doesn't specify how and where the commitments need to be stored and leaves it to schools to provide their feedback.
-Potential solutions include a centralized timestamping server, a permissioned blockchain with the different schools as the nodes and simple timestamping commitments to an existing blockchain.
+Potential solutions include a centralized time-stamping server, a permissioned blockchain with the different schools as the nodes and simple time-stamping commitments to an existing blockchain.
 All three options have their benefits and constraints and Kratos would arrive at a final recommendation depending on the option the majority of schools are convenient with
@@
 Building a common script for all the vendors poses a challenge since the number of vendors and the different fields and standards they follow are ever changing.
-This templated script would be trigerred automatically by Kratos each time it receives an incoming report and the report itself will be parsed ot udnerstand which vendor it originates from.
+This templated script would be triggered automatically by Kratos each time it receives an incoming report and the report itself will be parsed to understand which vendor it originates from.
 
@@
     \label{sec:15}
-    \item Zeide, E. (2016). 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?.
+    \item Zeide, E. (2016). 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?
+
+    Project Unicorn.
+
 \end{enumerate}
+\begin{figure}
+    \centering
+    \includegraphics[width=15cm]{{"images/Kratos.jpg}}
+    \caption{Kratos Architecture}
+    \label{fig:my_label}
+\end{figure}
 \end{document}

### 2019-06-04 paper.tex
update with vela's changes

@@
 Third, while data helps teachers improve pedagogical practices, an increasingly data-driven decision-making process suggests that student dimensions of learning and equitable participation in curriculum design becomes secondary.
-Finding a balance between increasingly data-driven decision making and student voice and choice is critical for an efficient educational ecosystem.
+Finding a balance between data-driven decision making and student voice is critical for an efficient educational ecosystem.
 In this paper, we introduce Kratos: an immutable decentralised data management system that provides data privacy and applied data literacy while empowering students with a user interface for data governance and active participation in the educational ecosystem.
-Using the advantages afforded by blockchain technologies, Kratos provides easy authentication, confidentiality, and accountability for data access, use, and sharing.
-The objective of Kratos is thus to equip schools and students with the ability to access, manage, control, and understand data and how, why, and by whom data is accessed and used at any one point in time without compromising student agency and privacy.
-This paper describes proof of concept for a decentralised data management system and its benefits to the educational ecosystem.
+Using the advantages of blockchain technologies, Kratos enables easy authentication and access to data.
+The objective of Kratos is thus to equip schools and students with the ability to access, manage and control their data and to understand how, why, and by whom data is accessed without compromising student agency and privacy.
+This paper describes proof of concept for Kratos and its benefits to the educational ecosystem.
 \bigbreak
@@
 \bigbreak
-On the other hand, concerns arise about the pervasiveness of fine-grained data collected about students [Zeide, 2017] and the risk of creating a 'permanent record' that can impede upon learners' futures [Cody, 2013].
-The use of online platforms, programs, and applications in schools often provided by for-profit vendors, many of whom have unclear policies about data privacy and third-party sharing [Common Sense Education, 2018], generates continuous stream of data about students' behavior and performance at an unprecedented scale [Zeide, 2017].
-Big data complicates traditional understanding of what constitutes sensitive information and what information serves an educational purpose.
-Data continues to drive decision-making of practitioners, vendors, and policy makers, marginalizing student dimensions of learning and equitable participation in the curriculum design.
-\bigbreak
-Interoperability challenges between vendors and schools at local, district, state, or federal level further pose barriers not only to cohesive data management and sharing but also lacks technological privacy infrastructure and accountability [Zeide, 2014; Common Sense Media Research, 2018].
-While digitizing and collecting student data is not new [Fitzgerald, 2014], technological developments like cloud computing and the growing use of web applications, learning management systems that gather fine-grained student data, amplify the concerns about data transfer, storage, use, and analysis [Sultan, 2010].
-\bigbreak
 Increasingly digitized learning environments enable constant data collection and algorithm-based assessment which change core school functions of teaching, assessment, and accreditation [Zeide, 2017].
 Digital systems for teaching and assessment drive data-based decision-making and enable micromanagement of students that can further pose restrictions on teacher autonomy and student (and parent) participation and willingness to challenge education decision-making.
-Interactive learning environments allow students to manage coursework online, take tests, and use a variety of online learning material.
-Schools deploy management systems for sorting and storing reporting and operational data.
-While some data is stored on local servers at school district level, other school data is managed by third-party vendors [Zhang, Li and Hao, 2015], who may consider data proprietary and thus inaccessible.
-This puts additional barriers to effective data sharing and comprehensive understanding of school data.
-\bigbreak
-The present challenges of fragmented data access and use, lack of student agency, accountability, and auditability to data access and use demand a socio-technological solution, one which can ensure that educational data can be used to help improve school processes while student privacy, agency, and future opportunities are not diminished.
+Schools deploy management systems for sorting and storing data and while some data is stored by schools at the district level, other data is managed by third-party vendors [Zhang, Li and Hao, 2015] who may consider this data proprietary and thus inaccessible.
+This puts barriers to effective data sharing and comprehensive understanding of school data.
 \bigbreak
@@
 Where teachers obtain data it requires further work to convert it into meaningful information upon which the teacher can adjust and plan instruction (Data Quality Campaign, 2018).
-  \item Unified data standardization challenges - The different vendors that CPS partners have their own implementation, most of them not regulated by any standards.
-The schools themselves follow a common standard defined by the SIF but there is no way to enforce this upon vendors.
-Even if the vendor agreed to follow a standard defined by the SIF, it is not possible to concretely verify this claim.
-  \item Lack of transparency - The vendors do not provide a comprehensive list of the data they collect from students.
-This makes data audits extremely difficult since we don't exhaustively know what to audit.
-  \item Security - It is impossible to verify if vendors are following best practices in data storage and retrieval.
-This is extremely dangerous since a malicious entity can arbitrarily change or retrieve student records from a poorly maintained database
+
+  \item Lack of data standard compliance - School data frameworks vary across districts and states.
+Additionally, vendors providing education technologies use different formats, data schema, and elements to organise and store student data as no single standard can be enforced upon them to comply with.
+Thus, disparate data scattered across different systems with varying degrees of accessibility and usability poses challenges to educators and other stakeholders in the education sector in assessing the impact of education technologies on the learning process.
+Moreover, lack of data standard compliance further limits the potential for imposing privacy law as is the case with the General Data Protection Regulation (GDPR), which provides sound privacy practices that are, on the one hand, enabled by data interoperability, and on the other, imposing vendors to maintain data transparency.
+
+  \item Lack of transparency - The scale, complexity, and the number of providers pose challenges for schools to have a comprehensive list of the data that is generated about and by students.
+Lack of data interoperability standard, which would simply the process of compiling, organising, and documenting what information is generated about students, how this information is used and by whom, makes data auditability difficult, near impossible, to achieve, less so to be useful to the instruction and learning processes.
+
+  \item Security - Due to lack of transparency with regards to how various education technology vendors organise and structure their data, it becomes near impossible to understand the security that they ensure for the data their products and services generate.
+A recent case  presents the fears that exist with regards to school data (Cameron, 2017).
+Schoolzilla, a K-12 education data service provider, detected a data breach, which exposed the personal data of 1.3 million K-12 students.
+Some argue (Hand, 2018) that the current available cybersecurity protocols are inefficient to secure student data while at the same time allow its use for analytics without compromising any part of it.
+Having inadequate access to all available data further limits the potentials to benefit fully from it.
+
 \end{itemize}
@@
 
-\subsection{Student agency and participation}
-The UNESCO framework for educational planning states that "the concern of planners is twofold: to reach a better understanding of the validity of education in its own empirically observed specific dimensions and to help in defining appropriate strategies for change" (Haddad, 1995, pp.
-5-6).
-
-While summative and cumulative assessments provide "empirically observed specific dimensions" about student academic performance, student agency and active participation is equally required in order for policy and education to design "strategies for change".
-Growing use of education technologies enables data-driven decision-making [Gibson et al., 2015].
-Technology-mediated instruction and assessment tools with learning analytics functionality track and diagnose student progress.
-Most educational technologies can interpret and equip educators with information via digital dashboards and 'skill meters'[Baker, 2016; New, 2016].
-Fine-grained and continuously accumulated data about student behavior and performance surpasses traditional notions of assessment [Zeide, 2017] posing limitations over student dimensions of learning and equitable participation in the curriculum design.
-
-Our prototype provides a graphical interface for student involvement in and accessibility to school data.
-While students and equally their parents can become acquainted with any changes and meanings of school data, students can also participate with personal feedback to the learning process.
-Student participation with personal perspectives and reflection are integral to the learning process.
-Such reflection is invariably carried out through questionnaire surveys examining school climate [Holahan and Batey, 2019].
-Both school climate surveys and the proposed student feedback application provide flexible selection of questionnaires and measurements with a common goal to improve school climate.
-However, the proposed application enables not only feedback from older students [Ibid.] but from all students, provided that the feedback addresses the common goal.
-The application further enables student agency and control over the frequency of feedback.
-
-\subsection{Data literacy}
-Data literacy is the ability to understand, create, and communicate data as information.
-Beyond that, in an increasingly data-driven socio-economic systems it becomes imperative not only to understand data but interact with it and participate in the decision-making processes it increasingly exerts its influence.
-While data provides insight about how systems operate, the inferences drawn from it is without context.
-Fine-grained data collected over long periods of time can lead to negative impact on individual well-being and pose limitations over future opportunities [Gasser et al., 2018].
-The growing digitization of schools create learning environments that enable constant data generation, the access to and use of which becomes hard to control, audit, and account for.
-The risk of digitized school environments constantly generating data and algorithmic assessments becoming permanent record that at any one point in time may limit individual opportunities leads us to prioritise on developing a socio-technological solution that enables applied data literacy for students at a young age.
-\textit{a note on Parents for Privacy Consortium and Data Quality Standards who make such efforts to educate students and parents about school data and that we (Kratos) complement these efforts.}
-
+Interoperability challenges between vendors and schools at local, district, state, or federal level further pose barriers not only to cohesive data management and sharing but also lacks technological privacy infrastructure and accountability [Zeide, 2014; Common Sense Media Research, 2018].
+While digitizing and collecting student data is not new [Fitzgerald, 2014], technological developments like cloud computing and the growing use of web applications, learning management systems that gather fine-grained student data, amplify the concerns about data transfer, storage, use, and analysis [Sultan, 2010].
+\bigbreak
+Concerns about the over-prevalence of fine-grained data collected about students [Zeide, 2017] and the risk of creating a 'permanent record' that can impede upon learners' futures [Cody, 2013] arise.
+The use of online platforms and applications in schools often provided by for-profit vendors, many of whom have unclear policies about data privacy and third-party sharing [Common Sense Education, 2018], generates continuous stream of data about students' behavior and performance at an unprecedented scale [Zeide, 2017].
+Big data complicates traditional understanding of what constitutes sensitive information and what information serves an educational purpose.
+Data however, continues to drive decision-making of practitioners and vendors, marginalizing student dimensions of learning and equitable participation in curriculum design.
+\bigbreak
+The present challenges of fragmented data access and use, lack of student agency, accountability, and auditability to data access and use demand a socio-technological solution, one which can ensure that educational data can be used to help improve school processes while student privacy, agency and future opportunities are not diminished.
 \section{Kratos}
-Kratos aims to define a set of principles and guidelines that schools should follow and provides a platform for both schools and students to see what kind of data is being collected by vendors along with ways to tweak data access.
+Kratos aims to define a set of principles and guidelines that schools should follow and provides a platform for both schools and students to see what kind of data is being collected by vendors along with ways to ensure that there is an immutable log and auditability for student data changes, access, and use.
 The goal for students would be to see how their data is being used and above the age of 18, gain access to this data.
-We define this as Data Ownership and Visibility- how students can own the data after the age of 18 and how students can see what their data is being used for.
+We define this as Data Ownership and Visibility - how students can own the data after the age of 18 and how students can see what their data is being used for.
 The goal for the school would be to make disparate data from different vendors compatible with each other and we define this problem as Data Interoperability.
 
-\subsection{Data ownership}
-In conventional systems, data ownership can be proved with the help of an access token but this provides no guarantee on when the owner came into possession of the data.
-In order to attest ownership at a specific point in time, we need time-stamping services like those described in \hyperref[sec:1]{[1]}, \hyperref[sec:2]{[2]} and \hyperref[sec:3]{[3]}.
-A time-stamping service requires something to be committed along with the timestamp and Kratos envisions this to be a cryptographic hash which also acts as an access token which vendors can use to access the data.
-\bigbreak
-The aim of using a cryptographic hash like SHA-3 \hyperref[sec:6]{[6]} is to ensure a uniquely random reference to the data to avoid out of channel data leaks.
-Kratos suggests using an element of randomness like a salt to generate the hash in order to have the ability to revoke tokens by regenerating randomness.
-Kratos enforces that all schools encrypt their data before creating access tokens to mitigate the risk of loss/theft of data.
-Past studies like \hyperref[sec:4]{[4]} and \hyperref[sec:5]{[5]} show that firms are willing to circumnavigate laws to collect data and encrypting data by design ensures that no third party can have access without being granted so explicitly.
-\bigbreak
-Kratos does not specify at which level data needs to be encrypted - it can be at the school level or at the student level depending on existing frameworks and rules surrounding the school.
-If the school chooses to encrypt student data on behalf of the student, Kratos enforces that the school use a unique key for each individual to minimize the risk of key theft.
-In addition to this, Kratos suggests that schools store their encrypted data in a distributed file storage system like IPFS to ensure data redundancy and availability in case of a mishap.
-Storing data on IPFS also makes it easier to create timestamps since it is sufficient to reference the IPFS pointer instead of potentially hashing the whole data.
-\bigbreak
-In the event a user wants to revoke access to a particular vendor, he could do so by changing the encryption key or by changing the randomness used to generate the access token.
-We suggest users don't regenerate encryption keys but Kratos does not enforce this and will provide end users with the option to choose between the two.
-\bigbreak
-Since schools have their own sets of policies, Kratos doesn't strictly enforce a set of practices for users to follow.
-This ensures that adoption of Kratos isn't constrained by a certain set of rules.
-At the same time, Kratos defines a set of minimum requirements to be on board to ensure good practices on date protection are followed.
-\bigbreak
-Kratos also doesn't specify how and where the commitments need to be stored and leaves it to schools to provide their feedback.
-Potential solutions include a centralized time-stamping server, a permissioned blockchain with the different schools as the nodes and simple time-stamping commitments to an existing blockchain.
-All three options have their benefits and constraints and Kratos would arrive at a final recommendation depending on the option the majority of schools are convenient with
-\bigbreak
-
+INSERT KRATOS.JPG HERE
 \subsection{Data Interoperability}
@@
 
+\subsection{Data ownership}
+In conventional systems, data ownership can be proved with the help of an access token but this provides no guarantee on when the owner came into possession of the data.
+In order to attest ownership at a specific point in time, we need time-stamping services like those described in \hyperref[sec:1]{[1]}, \hyperref[sec:2]{[2]} and \hyperref[sec:3]{[3]}.
+A time-stamping service requires something to be committed along with the timestamp and Kratos envisions this to be a cryptographic hash which also acts as an access token which vendors can use to access the data.
+\bigbreak
+The aim of using a cryptographic hash like SHA-3 \hyperref[sec:6]{[6]} is to ensure a uniquely random reference to the data to avoid out of channel data leaks.
+Kratos suggests using an element of randomness like a salt to generate the hash in order to have the ability to revoke tokens by regenerating randomness.
+Kratos enforces that all schools encrypt their data before creating access tokens to mitigate the risk of loss/theft of data.
+Past studies like \hyperref[sec:4]{[4]} and \hyperref[sec:5]{[5]} show that firms are willing to circumnavigate laws to collect data and encrypting data by design ensures that no third party can have access without being granted so explicitly.
+\bigbreak
+Kratos does not specify at which level data needs to be encrypted - it can be at the school level or at the student level depending on existing frameworks and rules surrounding the school.
+If the school chooses to encrypt student data on behalf of the student, Kratos enforces that the school use a unique key for each individual to minimize the risk of key theft.
+In addition to this, Kratos suggests that schools store their encrypted data in a distributed file storage system like IPFS to ensure data redundancy and availability in case of a mishap.
+Storing data on IPFS also makes it easier to create timestamps since it is sufficient to reference the IPFS pointer instead of potentially hashing the whole data.
+\bigbreak
+In the event a user wants to revoke access to a particular vendor, he could do so by changing the encryption key or by changing the randomness used to generate the access token.
+We suggest users don't regenerate encryption keys but Kratos does not enforce this and will provide end users with the option to choose between the two.
+\bigbreak
+Since schools have their own sets of policies, Kratos doesn't strictly enforce a set of practices for users to follow.
+This ensures that adoption of Kratos isn't constrained by a certain set of rules.
+At the same time, Kratos defines a set of minimum requirements to be on board to ensure good practices on date protection are followed.
+\bigbreak
+Kratos also does not specify how and where the commitments need to be stored and leaves it to schools to provide their feedback.
+Potential solutions include a centralized time-stamping server, a permissioned blockchain with the different schools as the nodes and simple time-stamping commitments to an existing blockchain.
+All three options have their benefits and constraints and Kratos would arrive at a final recommendation depending on what schools support best.
+\bigbreak
+
 \subsection{User interface}
 Visuals, wireframes, mockups, description of these
+
+\subsection{Student agency and participation}
+The UNESCO framework for educational planning states that "the concern of planners is twofold: to reach a better understanding of the validity of education in its own empirically observed specific dimensions and to help in defining appropriate strategies for change" (Haddad, 1995, pp.
+5-6).
+\bigbreak
+While summative and cumulative assessments provide "empirically observed specific dimensions" about student academic performance, student agency and active participation is equally required in order for policy and education to design "strategies for change".
+Growing use of education technologies enables data-driven decision-making [Gibson et al., 2015].
+Technology-mediated instruction and assessment tools with learning analytics functionality track and diagnose student progress.
+Most educational technologies can interpret and equip educators with information via digital dashboards and 'skill meters'[Baker, 2016; New, 2016].
+Fine-grained and continuously accumulated data about student behavior and performance surpasses traditional notions of assessment [Zeide, 2017] posing limitations over student dimensions of learning and equitable participation in the curriculum design.
+\bigbreak
+Our prototype provides a graphical interface for student involvement in and accessibility to school data.
+While students and equally their parents can become acquainted with any changes and meanings of school data, students can also participate with personal feedback to the learning process.
+Student participation with personal perspectives and reflection are integral to the learning process (Ackermann, 1996).
+Some reflection is invariably carried out through questionnaire surveys examining school climate (Holahan and Batey, 2019).
+Both school climate surveys and the proposed student feedback application provide flexible selection of questionnaires and measurements with a common goal to improve school climate.
+However, the proposed application enables not only feedback from older students (Ibid.) but from all students, provided that the feedback addresses the goal to encourage perspective-taking and reflection that are deemed necessary to the learning process (Kegan, 1982).
+The application further enables student agency and control over the frequency, depth, and nature of the feedback provided that it directly reflects the learning experience.
+
+\subsection{Data literacy}
+Data literacy is the ability to understand, create, and communicate data as information.
+Beyond that, in increasingly data-driven systems it becomes imperative not only to understand data but to interact with it and participate in the decision-making processes that it increasingly begins to impact.
+While data provides insight about how systems operate, the inferences drawn from it is without context.
+Fine-grained data collected over long periods of time can lead to negative impact on individual well-being and pose limitations over future opportunities [Gasser et al., 2018].
+The growing digitization of schools create learning environments that enable constant data generation, the access to and use of which becomes hard to control, audit, and account for.
+The risk of digitized school environments constantly generating data and algorithmic assessments becoming permanent record that at any one point in time may limit individual opportunities leads us to prioritise on developing a socio-technological solution that enables applied data literacy for students at a young age.
+\textit{a note on Parents for Privacy Consortium and Data Quality Standards who make such efforts to educate students and parents about school data and that we (Kratos) complement these efforts.}
 
@@
 
+Unlike other sectors such as retail (Lu and Xu, 2017), finance (Ito et al., 2017), and medicine (Acbo et al., 2019), education is still not truly leveraging from the digital transformation it is undergoing.
+Students generate large quantities of data from using various devices and applications in class.
+Schools generate large quantities of data about students.
+All of this data remains largely scattered across and owned by different, often for-profit, providers, many of whom consider their data proprietary.
+Scattered, disparate and unavailable data poses interoperability challenges, which creates further barriers to the generation of comprehensive data analysis - a necessary feedback for teachers and curriculum designers.
+Some companies provide data mining and user profiling as a result of combining a number of student data such as assessment grades, attendance, readiness surveys, demographics, and teacher assessment notes.
+However, these techniques come at a great cost while they are also challenged to meet acceptable standards for reliability and validity.
+Most crucially, student data privacy remains at the courtesy of such companies to protect it while students and schools do not gain any interest from their own data.
+
 - a few paragraphs on data and learning analytics..
 
-- paragraph on Kratos as an add-on, rather than replacement of existing efforts:
+\section{Acknowledgements}
+Rayner, Mario, Geoffrey?
 
@@
     \item Zeide, E. (2016). 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?
-
-    Project Unicorn.
+    \item Kegan, R. (1982). \textit{The evolving self: Problem and process in human development}. Cambridge: Harvard University Press.
+    \item Ackermann, E.
+(1996).
+Perspective-Taking and object Construction.
+\textit{In Constuctionism in Practice: Designing, Thinking, and Learning in a Digital World} (Kafai, Y., and Resnick, M., Eds.).
+Mahwah, New Jersey: Lawrence Erlbaum Associates.
+Part 1, Chap.
+2.
+pp.
+25-37.
+    \item Gibson D.C., Webb M.E., Ifenthaler D.
+(2019) Measurement Challenges of Interactive Educational Assessment.
+In: Sampson D., Spector J., Ifenthaler D., Isaías P., Sergis S.
+(eds) Learning Technologies for Transforming Large-Scale Teaching, Learning, and Assessment.
+Springer, Cham
 

### 2019-06-04 paper.tex
more changes

@@
 
-INSERT KRATOS.JPG HERE
+\textbf{INSERT KRATOS.JPG HERE}
+
 \subsection{Data Interoperability}
@@
 Kratos attempts to solve this issue by proposing a solution where different fields described by different vendors can be mapped to a single underlying scheme defined by Kratos and third party vendors would not have to worry about using disparate fields.
+
+\textbf{INSERT DATA ARCHITECTURE.JPG HERE}
 
@@
 Common examples of buckets would be PII (Personally Identifiable Information), Grade Reports (containing grade reports for each subject) and Attendance Records (recording the class wise attendance of the student).
-More such buckets will be added after receiving input from students and teachers on what the best ways to classify this data are.
+To formalise the type of data buckets it is imperative to develop a data taxonomy that comprises the existing school data standards and data elements according to which different education technology providers collect and organise their data.
 
@@
 The models and scripts developed as part of enabling interoperability will be open source and subject to continuous updates.
-We would also be having a complete code audit before relasing the system in a production environment to ensure the model performs and behaves the way it was designed and intended to.
+We would also be having a complete code audit before releasing the system in a production environment to ensure the model performs and behaves the way it was designed and intended to.
 
 \bigbreak
-The buckets defined by Kratos can be used in multiple ways: students can have a better understanding of how their data is being used, parents can know hwat kind of data is being collected by their children and administrators can ensure that sensitive information is not being shared with third party vendors.
-After the age of 18, this control would be given to the hands of students and they can decide if they wish to continue sharing data with these parties.
+The buckets defined by Kratos can be used in multiple ways: students can have a better understanding of how their data is being used, parents can know what kind of data is being collected about their children, and administrators can ensure that sensitive information is not being shared with third-party vendors.
+After legal age of consent, this control would be given to the hands of students and they can decide if they wish to continue sharing data with these parties.
 
-\subsection{Data ownership}
-In conventional systems, data ownership can be proved with the help of an access token but this provides no guarantee on when the owner came into possession of the data.
-In order to attest ownership at a specific point in time, we need time-stamping services like those described in \hyperref[sec:1]{[1]}, \hyperref[sec:2]{[2]} and \hyperref[sec:3]{[3]}.
-A time-stamping service requires something to be committed along with the timestamp and Kratos envisions this to be a cryptographic hash which also acts as an access token which vendors can use to access the data.
+\subsection{Data control}
+In conventional systems, data control or 'ownership' can be proved with the help of an access token but this provides no guarantee on when the owner came into possession of the data.
+In order to attest 'ownership' at a specific point in time, we need time-stamping services like those described in \hyperref[sec:1]{[1]}, \hyperref[sec:2]{[2]} and \hyperref[sec:3]{[3]}.
+A time-stamping service requires something to be committed along with the time-stamp and Kratos envisions this to be a cryptographic hash which also acts as an access token which vendors can use to access any student-generated data.
 \bigbreak
-The aim of using a cryptographic hash like SHA-3 \hyperref[sec:6]{[6]} is to ensure a uniquely random reference to the data to avoid out of channel data leaks.
+The aim of using a cryptographic hash like SHA-3 \hyperref[sec:6]{[6]} is to ensure a uniquely random reference to the data to avoid out-of-channel data leaks.
 Kratos suggests using an element of randomness like a salt to generate the hash in order to have the ability to revoke tokens by regenerating randomness.
@@
 If the school chooses to encrypt student data on behalf of the student, Kratos enforces that the school use a unique key for each individual to minimize the risk of key theft.
-In addition to this, Kratos suggests that schools store their encrypted data in a distributed file storage system like IPFS to ensure data redundancy and availability in case of a mishap.
+In addition to this, Kratos suggests that schools store their encrypted data in a distributed file storage system like IPFS to ensure data redundancy and availability in case of a setback.
 Storing data on IPFS also makes it easier to create timestamps since it is sufficient to reference the IPFS pointer instead of potentially hashing the whole data.
@@
 In the event a user wants to revoke access to a particular vendor, he could do so by changing the encryption key or by changing the randomness used to generate the access token.
-We suggest users don't regenerate encryption keys but Kratos does not enforce this and will provide end users with the option to choose between the two.
+We suggest users do not regenerate encryption keys but Kratos does not enforce this and will provide end users with the option to choose between the two.
 \bigbreak
-Since schools have their own sets of policies, Kratos doesn't strictly enforce a set of practices for users to follow.
-This ensures that adoption of Kratos isn't constrained by a certain set of rules.
-At the same time, Kratos defines a set of minimum requirements to be on board to ensure good practices on date protection are followed.
+Since schools have their own sets of policies, Kratos does not strictly enforce a set of practices for users to follow.
+This ensures that adoption of Kratos is not constrained by a certain set of rules.
+At the same time, Kratos defines a set of minimum requirements to be on board to ensure good practices on data protection are followed.
 \bigbreak
@@
 \bigbreak
-
-\subsection{User interface}
-Visuals, wireframes, mockups, description of these
 
@@
 \section{Discussion and future work}
-In this paper we propose Kratos, a data management system, which enables data interoperability for otherwise disparate data, applied data literacy, student agency over their data and participation in designing their education.
+In this paper we propose Kratos, a decentralised data management system, which enables data interoperability, student agency, applied data literacy, and student participation in the curriculum design.
+This proposal envisions an ideal techno-social solution that empowers the individual student by providing more control over their school data through data transparency, accountability, and auditability.
+At the same time, continuous legal (GDPR, 2016; Colorado Student Data Transparency and Security Act, 2016; Kelly, 2019) and policy (Student Data Privacy Consortium, 2018) efforts are being made in the right direction that will further help establish Kratos as a tool integral to the school ecosystem.
 
-Unlike other sectors such as retail (Lu and Xu, 2017), finance (Ito et al., 2017), and medicine (Acbo et al., 2019), education is still not truly leveraging from the digital transformation it is undergoing.
-Students generate large quantities of data from using various devices and applications in class.
-Schools generate large quantities of data about students.
-All of this data remains largely scattered across and owned by different, often for-profit, providers, many of whom consider their data proprietary.
-Scattered, disparate and unavailable data poses interoperability challenges, which creates further barriers to the generation of comprehensive data analysis - a necessary feedback for teachers and curriculum designers.
-Some companies provide data mining and user profiling as a result of combining a number of student data such as assessment grades, attendance, readiness surveys, demographics, and teacher assessment notes.
-However, these techniques come at a great cost while they are also challenged to meet acceptable standards for reliability and validity.
-Most crucially, student data privacy remains at the courtesy of such companies to protect it while students and schools do not gain any interest from their own data.
+To leverage decentralisation of a large scale data management system, Kratos shows that as learning analytics are crucial for educators to improve on instruction and pedagogy, the platform prioritises student agency, privacy, and control over their data.
 
-- a few paragraphs on data and learning analytics..
+Unlike other sectors such as retail (Lu and Xu, 2017), finance (Ito et al., 2017), and medicine (Acbo et al., 2019), education is still not fully leveraging from the digital transformation it is undergoing.
+In medicine, for instance, while data interoperability is important in order for various stakeholders to access comprehensive patient records, decentralised technologies have been used to enable such access while prioritising patient agency, privacy, and security (Agbo et al., 2019).
 
-\section{Acknowledgements}
-Rayner, Mario, Geoffrey?
+On the other hand, while data analytics provide insight about school processes, not everything that is measured is meaningful just as not everything meaningful can be given a numeric value.
+Students’ perspectives and reflection should equally count next to data-driven decision-making.
+Moreover, the sophistication of data analytics and long-term data collections further increase the risks of individual privacy.
+The risk of being publicly exposed can lead to the risk of being embarrassed or discriminated (Altman et al.
+2018).
+Being aware of such risks can diminish one’s sense of freedom for self-expression and in the case with children and young people - the freedom to simply try things out and make mistakes.
+Therefore, a decentralised data management system that prioritises student agency, privacy, and control over data, becomes crucial to the development of a safe space in which learners can learn and voice their perspectives without fear.
+
+Future work includes three distinctive steps.
+First, we plan to develop a comprehensive taxonomy of school data.
+This will serve to develop a comprehensive data vocabulary for applied data literacy and identify data elements.
+The second step involves developing prototype and formalising platform sections and functionality.
+And step three involves carrying out user studies to examine student experience, develop and fine-tune UI.
+
+\section{Acknowledgement}
+We thank Cambridge Public School district, Student Data Privacy Consortium, Acess for Learning (SIF) for their support.
+We thank Rayner Ng Jing Kai from Yale OpenLab (NUS) for the mockup designs, Mario Galea from  Geoffrey?
 
@@
 (eds) Learning Technologies for Transforming Large-Scale Teaching, Learning, and Assessment.
-Springer, Cham
+Springer, Cham.
+    \item Agbo, C. , Mahmoud, Q., and Eklund, J. (2019). Blockchain technology in healthcare: A systematic review. \textit{Healthcare, 7\textit{ (56)}}, 1-30.
+    \item Lu, Q. and Xu, X. (2017). Adaptable blockchain-based systems: A case study for product traceability. \textit{IEEE Software, 34\textit{ (6)}}, 21-27.
+    \item Ito, J., Narula, N., and Ali, R.
+(2017).
+The blockchain will do to the financial system what the Internet did to the media.
+\textit{Harvard Business Review}, March 23.
+Available from: https://hbr.org/2017/03/the-blockchain-will-do-to-banks-and-law-firms-what-the-internet-did-to-media
+    \item U.S.
+Department of Education, Office of Educational Technology (2012).
+\textit{Enhancing teaching and learning through educational data mining and learning analytics: An issue brief}.
+U.S.
+Department of education: Washington, D.C.
+    \item EU General Data Protection Regulation (GDPR): Regulation (EU) 2016/679 of the European Parliament and of the Council of 27 April 2016 on the protection of natural persons with regard to the processing of personal data and on the free movement of such data, and repealing Directive 95/46/EC (General Data Protection Regulation), OJ 2016 L 119/1.
+    \item Kelly, M. (2019). New privacy bill would give parents an ‘Eraser Button’ and ban ads targeting children. \textit{The Verge}, March 12.
 

### 2019-06-05 paper.tex
add images

@@
 \usepackage{geometry}
+\usepackage{float}
 \usepackage[final]{graphicx}
@@
 To contextualise the complexity related to school data - the opportunities and challenges schools face as a result of the growing digitisation of operation and academic processes - Kratos has partnered with  Cambridge Public School (CPS) district that administer public elementary and high schools in Cambridge, Massachusetts, the Access for Learning (SIF), and the Student Data Privacy Consortium.
-With the help of CPS's Information and Communication Technology Services Chief Information Officer and Database Administrator, we explored data interoperability issues with them along with feedback on potential mitigation schemes:
+With the help of CPS' Information and Communication Technology Services Chief Information Officer and Database Administrator, we explored data interoperability issues with them along with feedback on potential mitigation schemes:
 
 \begin{itemize}
-  \item Data access - CPS is in agreement with over 100 vendors providing education technologies ([7]).
-The majority of these would mostly provide no direct and comprehensive access to data generated about and by students who make use of their products or services.
+  \item \textbf{Data access} - CPS is in agreement with over 100 vendors providing education technologies ([7]).
+A majority of these provide no direct and comprehensive access to data generated about and by students who use of their products.
 Where available, data access is  provided in the form of reports - a summary of information which the district database administrator can request and download.
-Some providers are even less flexible and only offer to give reports once in 45 days making it impossible for teachers to act on feedback that they might infer from the data.
 On other occasions vendors supply data directly to teachers as well as the students in the form of digital dashboards.
@@
 
-  \item Lack of data standard compliance - School data frameworks vary across districts and states.
-Additionally, vendors providing education technologies use different formats, data schema, and elements to organise and store student data as no single standard can be enforced upon them to comply with.
+  \item \textbf{Lack of data standard compliance} - School data frameworks vary across districts and states.
+Additionally, vendors providing education technologies use different data formats, schema and elements to organise and store student data as no single standard can be enforced upon them to comply with.
 Thus, disparate data scattered across different systems with varying degrees of accessibility and usability poses challenges to educators and other stakeholders in the education sector in assessing the impact of education technologies on the learning process.
-Moreover, lack of data standard compliance further limits the potential for imposing privacy law as is the case with the General Data Protection Regulation (GDPR), which provides sound privacy practices that are, on the one hand, enabled by data interoperability, and on the other, imposing vendors to maintain data transparency.
+Moreover, the lack of data standard compliance further limits the potential for imposing privacy law as is the case with the General Data Protection Regulation (GDPR), which provides sound privacy practices that are, on the one hand, enabled by data interoperability, and on the other, imposing vendors to maintain data transparency.
 
-  \item Lack of transparency - The scale, complexity, and the number of providers pose challenges for schools to have a comprehensive list of the data that is generated about and by students.
-Lack of data interoperability standard, which would simply the process of compiling, organising, and documenting what information is generated about students, how this information is used and by whom, makes data auditability difficult, near impossible, to achieve, less so to be useful to the instruction and learning processes.
+  \item \textbf{Lack of transparency} - The scale, complexity, and number of providers pose challenges for schools to have a comprehensive list of the data that is generated about students.
+Lack of data interoperability standards which would simply be the process of compiling, organising and documenting what information is generated about students makes data auditability difficult to achieve, less so to be useful to the instruction and learning processes.
 
-  \item Security - Due to lack of transparency with regards to how various education technology vendors organise and structure their data, it becomes near impossible to understand the security that they ensure for the data their products and services generate.
-A recent case  presents the fears that exist with regards to school data (Cameron, 2017).
-Schoolzilla, a K-12 education data service provider, detected a data breach, which exposed the personal data of 1.3 million K-12 students.
-Some argue (Hand, 2018) that the current available cybersecurity protocols are inefficient to secure student data while at the same time allow its use for analytics without compromising any part of it.
-Having inadequate access to all available data further limits the potentials to benefit fully from it.
+  \item \textbf{Security} - Due to lack of transparency with regards to how various education technology vendors organise and structure their data, it becomes difficult to understand the security they ensure for the data their products and services generate.
+A recent case  presents the fears that exist with regards to school data (Cameron, 2017) - Schoolzilla, a K-12 education data service provider, detected a data breach which exposed the personal data of 1.3 million K-12 students.
+Some argue (Hand, 2018) that currently available cybersecurity protocols are inefficient to secure student data while at the same time allowing data analytics.
+Having inadequate access to available data further limits potentials to fully benefit from it.
 
@@
 
-We can thus see that schools are finding it increasingly difficult to obtain data about their own students and are becoming isolated due to a lack of knowledge on how vendors are processing this data.
+We thus infer that schools are finding it increasingly difficult to obtain data about their own students and are becoming isolated due to a lack of knowledge on how vendors are processing this data.
 
@@
 
-Interoperability challenges between vendors and schools at local, district, state, or federal level further pose barriers not only to cohesive data management and sharing but also lacks technological privacy infrastructure and accountability [Zeide, 2014; Common Sense Media Research, 2018].
-While digitizing and collecting student data is not new [Fitzgerald, 2014], technological developments like cloud computing and the growing use of web applications, learning management systems that gather fine-grained student data, amplify the concerns about data transfer, storage, use, and analysis [Sultan, 2010].
+Interoperability challenges between vendors and schools at local, district, state, or federal level pose barriers to cohesive data management and sharing while lacking basic technological privacy infrastructure and accountability [Zeide, 2014; Common Sense Media Research, 2018].
+While digitizing and collecting student data is not new [Fitzgerald, 2014], technological developments like cloud computing and the Internet of Things (IoT) amplify concerns about data transfer, storage, use, and analysis [Sultan, 2010].
 \bigbreak
-Concerns about the over-prevalence of fine-grained data collected about students [Zeide, 2017] and the risk of creating a 'permanent record' that can impede upon learners' futures [Cody, 2013] arise.
-The use of online platforms and applications in schools often provided by for-profit vendors, many of whom have unclear policies about data privacy and third-party sharing [Common Sense Education, 2018], generates continuous stream of data about students' behavior and performance at an unprecedented scale [Zeide, 2017].
+Concerns arise about over-prevalence of fine-grained data collected about students [Zeide, 2017] and the risk of creating a 'permanent record' that can impede upon learners' futures [Cody, 2013].
+The use of online platforms and applications in schools, often provided by for-profit vendors many of whom have unclear policies about data privacy [Common Sense Education, 2018], generates a continuous stream of data about students' behavior and performance at an unprecedented scale [Zeide, 2017].
 Big data complicates traditional understanding of what constitutes sensitive information and what information serves an educational purpose.
-Data however, continues to drive decision-making of practitioners and vendors, marginalizing student dimensions of learning and equitable participation in curriculum design.
+Such data however, continues to drive decision-making of practitioners and vendors, marginalizing student dimensions of learning and equitable participation in curriculum design.
 \bigbreak
-The present challenges of fragmented data access and use, lack of student agency, accountability, and auditability to data access and use demand a socio-technological solution, one which can ensure that educational data can be used to help improve school processes while student privacy, agency and future opportunities are not diminished.
+The present challenges of fragmented data access, the lack of student agency and auditability and the lack of a concrete framework for data accessibility demand a socio-technological solution which can ensure that educational data can be used to help improve school processes while not diminishing student privacy, agency and future opportunities.
+
 \section{Kratos}
-Kratos aims to define a set of principles and guidelines that schools should follow and provides a platform for both schools and students to see what kind of data is being collected by vendors along with ways to ensure that there is an immutable log and auditability for student data changes, access, and use.
+Kratos aims to define a set of guidelines and principles that schools should follow and provides a platform for both schools and students to see what kind of data is being collected by vendors with the help of an immutable log for auditing student data changes, access, and use.
 The goal for students would be to see how their data is being used and above the age of 18, gain access to this data.
@@
 
-\textbf{INSERT KRATOS.JPG HERE}
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=15cm]{{"images/Kratos.jpg}}
+    \caption{Kratos Architecture}
+    \label{fig:my_label}
+\end{figure}
 
@@
 
-\textbf{INSERT DATA ARCHITECTURE.JPG HERE}
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=15cm]{{"images/data-architecture.jpg}}
+    \caption{Data Architecture}
+    \label{fig:my_label}
+\end{figure}
 
@@
 \bigbreak
-Kratos does not specify at which level data needs to be encrypted - it can be at the school level or at the student level depending on existing frameworks and rules surrounding the school.
+Kratos suggests that encryption be done at the student level but this can also be at the school level depending on existing frameworks and rules surrounding the school.
 If the school chooses to encrypt student data on behalf of the student, Kratos enforces that the school use a unique key for each individual to minimize the risk of key theft.
@@
 In the event a user wants to revoke access to a particular vendor, he could do so by changing the encryption key or by changing the randomness used to generate the access token.
-We suggest users do not regenerate encryption keys but Kratos does not enforce this and will provide end users with the option to choose between the two.
+We suggest users do not regenerate encryption keys but Kratos will provide users an option to choose between the two.
 \bigbreak
@@
 \bigbreak
-Kratos also does not specify how and where the commitments need to be stored and leaves it to schools to provide their feedback.
-Potential solutions include a centralized time-stamping server, a permissioned blockchain with the different schools as the nodes and simple time-stamping commitments to an existing blockchain.
-All three options have their benefits and constraints and Kratos would arrive at a final recommendation depending on what schools support best.
+Kratos also does not enforce how commitments need to be generated and stored, leaving it as an option for schools to provide their feedback on.
+Solutions provided by Kratos would include a centralized time-stamping server, a permissioned blockchain with the different schools as the nodes and simple time-stamping commitments to an existing blockchain.
+All three have their benefits and constraints and Kratos would enable schools to customize this to their requirements.
 \bigbreak
@@
 
+\begin{figure}[H]
+    \centering
+      \begin{minipage}[b]{0.4\textwidth}
+    \includegraphics[width=5cm]{{"images/home.png}}
+    \caption{Data Information}
+    \label{fig:my_label}
+    \end{minipage}
+    \begin{minipage}[b]{0.4\textwidth}
+    \includegraphics[width=5cm]{{"images/summary.png}}
+    \caption{Detailed info about fields}
+    \label{fig:my_label}
+    \end{minipage}
+\end{figure}
+
 \subsection{Data literacy}
@@
 
+\begin{figure}[H]
+    \centering
+      \begin{minipage}[b]{0.4\textwidth}
+    \includegraphics[height=9.5cm]{{"images/data-detail-1.png}}
+    \caption{Data Information}
+    \label{fig:my_label}
+    \end{minipage}
+      \begin{minipage}[b]{0.4\textwidth}
+    \includegraphics[height=9.5cm]{{"images/data-detail-2.png}}
+    \caption{Detailed info about fields}
+    \label{fig:my_label}
+    \end{minipage}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+      \begin{minipage}[b]{0.4\textwidth}
+    \includegraphics[height=9.5cm]{{"images/detail.png}}
+    \caption{Permit / Revoke Access}
+    \label{Permit / Revoke Access}
+    \end{minipage}
+      \begin{minipage}[b]{0.4\textwidth}
+    \includegraphics[height=9.5cm]{{"images/feedback-1.png}}
+    \caption{Feedback}
+    \label{fig:my_label}
+    \end{minipage}
+\end{figure}
+
 \section{Discussion and future work}
-In this paper we propose Kratos, a decentralised data management system, which enables data interoperability, student agency, applied data literacy, and student participation in the curriculum design.
-This proposal envisions an ideal techno-social solution that empowers the individual student by providing more control over their school data through data transparency, accountability, and auditability.
-At the same time, continuous legal (GDPR, 2016; Colorado Student Data Transparency and Security Act, 2016; Kelly, 2019) and policy (Student Data Privacy Consortium, 2018) efforts are being made in the right direction that will further help establish Kratos as a tool integral to the school ecosystem.
-
-To leverage decentralisation of a large scale data management system, Kratos shows that as learning analytics are crucial for educators to improve on instruction and pedagogy, the platform prioritises student agency, privacy, and control over their data.
-
-Unlike other sectors such as retail (Lu and Xu, 2017), finance (Ito et al., 2017), and medicine (Acbo et al., 2019), education is still not fully leveraging from the digital transformation it is undergoing.
-In medicine, for instance, while data interoperability is important in order for various stakeholders to access comprehensive patient records, decentralised technologies have been used to enable such access while prioritising patient agency, privacy, and security (Agbo et al., 2019).
-
-On the other hand, while data analytics provide insight about school processes, not everything that is measured is meaningful just as not everything meaningful can be given a numeric value.
+In this paper we propose Kratos, a decentralised data management system which enables data interoperability, applied data literacy, and student participation in the curriculum design.
+This proposal envisions an ideal techno-social solution that empowers students by providing more control over their school data through data transparency, accountability and auditability.
+At the same time, continuous legal (GDPR, 2016; Colorado Student Data Transparency and Security Act, 2016; Kelly, 2019) and policy (Student Data Privacy Consortium, 2018) efforts are being made that will further help establish Kratos as a tool integral to the school ecosystem.
+\bigbreak
+Unlike other sectors such as retail (Lu and Xu, 2017), finance (Ito et al., 2017), and medicine (Acbo et al., 2019), education is still not fully leveraging the power of global digital transformation.
+In medicine, for instance, data interoperability is important in order for various stakeholders to access comprehensive patient records and decentralised technologies have been used to enable such access while prioritising patient agency, privacy and security (Acbo et al., 2019).
+\bigbreak
+On the other hand, while data analytics provides insight about school processes, not everything that is measured is meaningful.
 Students’ perspectives and reflection should equally count next to data-driven decision-making.
-Moreover, the sophistication of data analytics and long-term data collections further increase the risks of individual privacy.
-The risk of being publicly exposed can lead to the risk of being embarrassed or discriminated (Altman et al.
+Moreover, the sophistication of data analytics and long-term data collection further decrease an individual's privacy.
+The risk of being publicly exposed can lead to being embarrassed or discriminated (Altman et al.
 2018).
-Being aware of such risks can diminish one’s sense of freedom for self-expression and in the case with children and young people - the freedom to simply try things out and make mistakes.
-Therefore, a decentralised data management system that prioritises student agency, privacy, and control over data, becomes crucial to the development of a safe space in which learners can learn and voice their perspectives without fear.
-
+Constantly being aware of such risks can diminish one’s sense of freedom for self-expression and in the case with children and young people - the freedom to try out new things and make mistakes.
+Therefore, a decentralised data management system that prioritises student privacy and control over data, becomes crucial to the development of a safe and free learning environment.
+\bigbreak
 Future work includes three distinctive steps.
@@
 This will serve to develop a comprehensive data vocabulary for applied data literacy and identify data elements.
-The second step involves developing prototype and formalising platform sections and functionality.
-And step three involves carrying out user studies to examine student experience, develop and fine-tune UI.
+The second step involves prototyping and formalising platform sections and functionality.
+And the third involves carrying out user studies to get student feedback and the fine tuning of the platform's user interface.
 
-\section{Acknowledgement}
+\section{Acknowledgements}
 We thank Cambridge Public School district, Student Data Privacy Consortium, Acess for Learning (SIF) for their support.
-We thank Rayner Ng Jing Kai from Yale OpenLab (NUS) for the mockup designs, Mario Galea from  Geoffrey?
+We would also like to thank Rayner Ng Jing Kai and Goeffrey Martin from Yale-NUS College for their help in designing mockup designs and giving feedback on data interoperability.
 
@@
 \end{enumerate}
-\begin{figure}
-    \centering
-    \includegraphics[width=15cm]{{"images/Kratos.jpg}}
-    \caption{Kratos Architecture}
-    \label{fig:my_label}
-\end{figure}
+
 \end{document}

### 2019-06-06 paper.tex
add veli edits + make all references APA

@@
 
-\title{Kratos: blockchain solution for data privacy, literacy, and student agency in a data-driven educational ecosystem}
-\author{Dr.\ Velislava Hillman \quad Varunram Ganesh \quad Geoffrey Martin}
-\date{May 2019}
+\title{Kratos: A solution for data privacy, literacy and student agency in a data-driven educational ecosystem}
+\author{Dr.\ Velislava Hillman \quad Varunram Ganesh}
+\date{June 2019}
 
@@
 \begin{itemize}
-  \item \textbf{Data access} - CPS is in agreement with over 100 vendors providing education technologies ([7]).
+  \item \textbf{Data access} - CPS is in agreement with over 100 vendors providing education technologies.
 A majority of these provide no direct and comprehensive access to data generated about and by students who use of their products.
@@
 Thus, disparate data scattered across different systems with varying degrees of accessibility and usability poses challenges to educators and other stakeholders in the education sector in assessing the impact of education technologies on the learning process.
-Moreover, the lack of data standard compliance further limits the potential for imposing privacy law as is the case with the General Data Protection Regulation (GDPR), which provides sound privacy practices that are, on the one hand, enabled by data interoperability, and on the other, imposing vendors to maintain data transparency.
+Moreover, the lack of data standard compliance further limits the potential for imposing privacy law as is the case with the General Data Protection Regulation (GDPR, 2016), which provides sound privacy practices that are, on the one hand, enabled by data interoperability, and on the other, imposing vendors to maintain data transparency.
 
@@
 
-We thus infer that schools are finding it increasingly difficult to obtain data about their own students and are becoming isolated due to a lack of knowledge on how vendors are processing this data.
+We thus infer that schools are finding it increasingly difficult to obtain data about their own students in a comprehensive and timely manner and are thus becoming isolated due to a lack of knowledge on how vendors are processing the data that is generated as a result of their products and services used in class.
 
@@
 
-Interoperability challenges between vendors and schools at local, district, state, or federal level pose barriers to cohesive data management and sharing while lacking basic technological privacy infrastructure and accountability [Zeide, 2014; Common Sense Media Research, 2018].
-While digitizing and collecting student data is not new [Fitzgerald, 2014], technological developments like cloud computing and the Internet of Things (IoT) amplify concerns about data transfer, storage, use, and analysis [Sultan, 2010].
+Interoperability challenges between vendors and schools at local, district, state, or federal level pose barriers to cohesive data management and sharing while lacking basic technological privacy infrastructure and accountability (Zeide, 2014; Common Sense Media Research, 2018).
+While digitizing and collecting student data is not new (Fitzgerald, 2014), technological developments like cloud computing and the Internet of Things (IoT) amplify concerns about data transfer, storage, use, and analysis (Sultan, 2010).
 \bigbreak
-Concerns arise about over-prevalence of fine-grained data collected about students [Zeide, 2017] and the risk of creating a 'permanent record' that can impede upon learners' futures [Cody, 2013].
-The use of online platforms and applications in schools, often provided by for-profit vendors many of whom have unclear policies about data privacy [Common Sense Education, 2018], generates a continuous stream of data about students' behavior and performance at an unprecedented scale [Zeide, 2017].
+Concerns arise about over-prevalence of fine-grained data collected about students (Zeide, 2017) and the risk of creating a 'permanent record' that can impede upon learners' futures (Cody, 2013).
+The use of online platforms and applications in schools, often provided by for-profit vendors many of whom have unclear policies about data privacy (Common Sense Education, 2018), generates a continuous stream of data about students' behavior and performance at an unprecedented scale (Zeide, 2017).
 Big data complicates traditional understanding of what constitutes sensitive information and what information serves an educational purpose.
@@
 However, work done by standardization bodies like SIF and CEDS has shown that this is at best partially effective since there is no financial or social incentive for vendors to migrate to the proposed set of standards.
-Kratos attempts to solve this issue by proposing a solution where different fields described by different vendors can be mapped to a single underlying scheme defined by Kratos and third party vendors would not have to worry about using disparate fields.
+Kratos attempts to solve this issue by proposing a solution where different fields described by different vendors can be mapped to a single underlying scheme defined by Kratos and third party vendors would not have to change their established data organisation and structure.
 
@@
     \includegraphics[width=5cm]{{"images/home.png}}
-    \caption{Data Information}
+    \caption{Student UI: Student view of data information}
     \label{fig:my_label}
@@
     \includegraphics[width=5cm]{{"images/summary.png}}
-    \caption{Detailed info about fields}
+    \caption{Student UI: Student view of detailed data fields}
     \label{fig:my_label}
@@
 \subsection{Data literacy}
-Data literacy is the ability to understand, create, and communicate data as information.
-Beyond that, in increasingly data-driven systems it becomes imperative not only to understand data but to interact with it and participate in the decision-making processes that it increasingly begins to impact.
+ Data literacy is still a debatable construct (Bowler and Acker, 2017).
+Data means quantified information, which is "situated, taking its meaning from its context and the perspective of its beholder" (Borgman, 2015, p.
+18).
+Making sense of data is relative.
+What data may mean to someone can mean nothing to others (Bowler and Acker, 2017).
+In increasingly data-driven systems it becomes imperative not only to understand data but to interact with it and participate in the decision-making processes that it increasingly begins to impact.
 While data provides insight about how systems operate, the inferences drawn from it is without context.
-Fine-grained data collected over long periods of time can lead to negative impact on individual well-being and pose limitations over future opportunities [Gasser et al., 2018].
-The growing digitization of schools create learning environments that enable constant data generation, the access to and use of which becomes hard to control, audit, and account for.
-The risk of digitized school environments constantly generating data and algorithmic assessments becoming permanent record that at any one point in time may limit individual opportunities leads us to prioritise on developing a socio-technological solution that enables applied data literacy for students at a young age.
-\textit{a note on Parents for Privacy Consortium and Data Quality Standards who make such efforts to educate students and parents about school data and that we (Kratos) complement these efforts.}
+Fine-grained data collected over long periods of time can lead to negative impact on individual well-being and pose limitations over future opportunities (Altman et al., 2018; Friedman and Nissenbaum, 1996).
+ \bigbreak
+ The growing digitization of schools create learning environments that enable constant data generation, the access to, and use of which becomes hard to control, audit, account for, and understand its impact.
+The risk from digitized school environments that constantly generate data into a permanent record (Sirota, 2013) that at any one point in time may limit individual opportunities leads us to prioritise on developing a techno-social solution that enables applied data literacy for students at an ever younger age.
+ \bigbreak
+ While some studies examine aspects of data literacy in young people's lives (Bowler and Acker, 2017; Selwyn and Pangrazio, 2018; Kumar et al., 2017), the focus remains on Internet and social media data.
+Studies related to young people's perceptions, knowledge, and understanding of school data literacy are meagre.
+ \bigbreak
+ Existing literature demonstrates that young people have varying interprations and a general understanding of data.
+However, many still find difficulties to connect with data at a concrete level, with the notion of having a data dossier (Bowler and Acker, 2017, p.
+27).
+This highlights the pressing need to develop strategies for school-related data literacy, introduce data management skills, and encourage student participation in an increasingly data-driven decision-making school environment.
+ \bigbreak
+ Kratos thus proposes a comprehensive, yet simple access, knowledge, and interaction with school data.
+At conceptualisation stage we choose three core data elements to work with.
+These are assessment, including test scores, reports, and credentialing; conduct, including observation reports, behaviour, and attendance; and personally identifiable information, including personal identifiers, demographics, parent information, medical reports, and socio-economic information.
+We take these three data elements crucial to student privacy and control.
+In our conceptual work (Figures 3-5) we recognise that while schools collect data for each of these three elements, digital applications used in class also collect similar such data.
+For example, as schools contain data about conduct, applications such as Class Dojo (Manolev et al., 2018) gather similar data.
+Teachers may further report behaviour using Swivl.
+Currently, students do not have a comprehensive access to or knowledge of all data that exists about their conduct.
+Moreover, there is no awareness and knowledge of the potential impact of all data that exists about their conduct.
+We therefore create visibility to the disparate data that exists about student conduct, provide visibility and control over who accesses such data and for what purpose, and enable learning about data (figure 6), its meaning and purpose.
 
@@
     \includegraphics[height=9.5cm]{{"images/data-detail-1.png}}
-    \caption{Data Information}
+    \caption{User UI: Student view of data}
     \label{fig:my_label}
@@
     \includegraphics[height=9.5cm]{{"images/data-detail-2.png}}
-    \caption{Detailed info about fields}
+    \caption{User UI: Student view of data and definition}
     \label{fig:my_label}
@@
 Unlike other sectors such as retail (Lu and Xu, 2017), finance (Ito et al., 2017), and medicine (Acbo et al., 2019), education is still not fully leveraging the power of global digital transformation.
-In medicine, for instance, data interoperability is important in order for various stakeholders to access comprehensive patient records and decentralised technologies have been used to enable such access while prioritising patient agency, privacy and security (Acbo et al., 2019).
+In medicine, for instance, data interoperability is important in order for various stakeholders to access comprehensive patient records and various cases across Europe (Demertzis et al., 2018) and the United States (Azaria et al., 2016) decentralised architectures have been developed to enable data sharing, interoperability, and access while prioritising patient agency, privacy, and security.
+Similarly, a more comprehensive data management for interoperability is highly needed in the education sector without compromising student privacy and control over their data.
 \bigbreak
@@
 \section{Acknowledgements}
-We thank Cambridge Public School district, Student Data Privacy Consortium, Acess for Learning (SIF) for their support.
-We would also like to thank Rayner Ng Jing Kai and Goeffrey Martin from Yale-NUS College for their help in designing mockup designs and giving feedback on data interoperability.
+We thank Cambridge Public School district, Student Data Privacy Consortium, Access for Learning (SIF) for their support.
+We would also like to thank Rayner Ng Jing Kai and Geoffrey Martin from Yale-NUS College for their help in sketching mock-up designs and giving feedback on data interoperability.
 
@@
     \label{sec:6}
-    \item Bertoni, G., Daemen, J., Peeters, M., Van Assche, G., \& Van Keer, R. (2012). Keccak implementation overview. URL  \url{http://keccak.neokeon.org/Keccak-implementation-3.2.pdf}.
+    \item Bertoni, G., Daemen, J., Peeters, M., Van Assche, G., \& Van Keer, R. (2012). Keccak implementation overview. URL  \url{http://keccak.neokeon.org/Keccak-implementation-3.2.pdf}
     \label{sec:7}
@@
     \label{sec:8}
-    \item Colins, A., and Halverson, R. (2018).\textit{ Re-thinking education in the age of technology: The digital revolution and schooling in America}. Teach Zeide, E. (2016).
+    \item Colins, A., and Halverson, R. (2018).\textit{ Re-thinking education in the age of technology: The digital revolution and schooling in America}. Teachers College Press: New York.
     \label{sec:9}
-    \item 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?.
+    Zeide, E. (2016). 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?.
     \label{sec:10}
-    \item Colins, A., and Halverson, R. (2018).\textit{ Re-thinking education in the age of technology: The digital revolution and schooling in America}. Teachers College Press: New York.
+    \item Cody, A. (2013). Will the data warehouse become every student and teacher's 'permanent record'? \textit{Education Week}, May, 20.
     \label{sec:11}
-    \item Cody, A. (2013). Will the data warehouse become every student and teacher's 'permanent record'? \textit{Education Week}, May, 20.
-    \label{sec:12}
     \item Gibson, D.
@@
 Challenges of big data in educational assessment.\textit{Proceedings of the IADIS International Conference of Exploratory Learning in Digital Age}, 92-100.
+    \label{sec:12}
+    \item Sultan, N. (2010). Cloud computing for education: a new dawn? 30 International Journal of Information Management  109.
     \label{sec:13}
-    \item Sultan, N. (2010). Cloud computing for education: a new dawn? 30 International Journal of Information Management  109.
+    \item Fitzgerald, B. (2014). Data collection isn't new. And it predates common core.Funny Monkey.
     \label{sec:14}
-    \item Fitzgerald, B. (2014). Data collection isn't new. And it predates common core.Funny Monkey.
+    \item Keegan, R. (1982). The evolving self: Problem and process in human development. Cambridge: Harvard UP.
     \label{sec:15}
-    \item Zeide, E. (2016). 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?
-    \item Kegan, R. (1982). \textit{The evolving self: Problem and process in human development}. Cambridge: Harvard University Press.
-    \item Ackermann, E.
-(1996).
-Perspective-Taking and object Construction.
-\textit{In Constuctionism in Practice: Designing, Thinking, and Learning in a Digital World} (Kafai, Y., and Resnick, M., Eds.).
-Mahwah, New Jersey: Lawrence Erlbaum Associates.
-Part 1, Chap.
-2.
-pp.
-25-37.
-    \item Gibson D.C., Webb M.E., Ifenthaler D.
-(2019) Measurement Challenges of Interactive Educational Assessment.
-In: Sampson D., Spector J., Ifenthaler D., Isaías P., Sergis S.
-(eds) Learning Technologies for Transforming Large-Scale Teaching, Learning, and Assessment.
+    \item Ackermann, E. (1996). \textit{Constructionism in practice: designing, thinking, and learning in a digital world}. Routledge.
+    \label{sec:16}
+    \item Gibson, D.
+C., Webb, M.
+E., & Ifenthaler, D.
+(2019).
+Measurement Challenges of Interactive Educational Assessment.
+In \textit{Learning Technologies for Transforming Large-Scale Teaching, Learning, and Assessment} (pp.
+19-33).
 Springer, Cham.
-    \item Agbo, C. , Mahmoud, Q., and Eklund, J. (2019). Blockchain technology in healthcare: A systematic review. \textit{Healthcare, 7\textit{ (56)}}, 1-30.
-    \item Lu, Q. and Xu, X. (2017). Adaptable blockchain-based systems: A case study for product traceability. \textit{IEEE Software, 34\textit{ (6)}}, 21-27.
-    \item Ito, J., Narula, N., and Ali, R.
+    \label{sec:17}
+    \item Agbo, C.
+C., Mahmoud, Q.
+H., & Eklund, J.
+M.
+(2019, June).
+Blockchain Technology in Healthcare: A Systematic Review.
+In \textit{Healthcare} (Vol.
+7, No.
+2, p.
+56).
+Multidisciplinary Digital Publishing Institute.
+    \label{sec:18}
+    \item Lu, Q., & Xu, X. (2017). Adaptable blockchain-based systems: a case study for product traceability. \textit{IEEE Software,} 34(6), 21-27.
+    \label{sec:19}
+    \item Ito, J., Narula, N., & Ali, R. (2017). The blockchain will do to the financial system what the internet did to media. \textit{Harvard Business Review,} 9(March).
+    \label{sec:20}
+    \item Bienkowski, M., Feng, M., & Means, B.
+(2012).
+Enhancing teaching and learning through educational data mining and learning analytics: An issue brief.
+\textit{US Department of Education, Office of Educational Technology,} 1, 1-57.
+    \label{sec:21}
+    \item Regulation, G.
+D.
+P.
+(2016).
+Regulation (EU) 2016/679 of the European Parliament and of the Council of 27 April 2016 on the protection of natural persons with regard to the processing of personal data and on the free movement of such data, and repealing Directive 95/46.
+\textit{Official Journal of the European Union (OJ)}, 59(1-88), 294.
+    \label{sec:22}
+    \item Kelly, M.
+New privacy bill would give parents an ‘Eraser Button’ and ban ads targeting children (2019).
+URL \url{https://www.theverge.com/2019/3/12/18261181/eraser-button-bill-children-privacy-coppa-hawley-markey}
+    \label{sec:23}
+    \item Demertzis, I., Papadopoulos, S., Papapetrou, O., Deligiannakis, A., Garofalakis, M., and Papamanthou, C.
+(2018).
+Practical Private Range Search in Depth.
+\textit{ACM Transactions on Database Systems, 43(1),} 2.
+    \label{sec:24}
+    \item Altman, M., Wood, A. B., O'Brien, D., & Gasser, U. (2018). Practical approaches to big data privacy over time.
+    \label{sec:25}
+    \item Bowler, L., Acker, A., Jeng, W., and Chi, Y.
 (2017).
-The blockchain will do to the financial system what the Internet did to the media.
-\textit{Harvard Business Review}, March 23.
-Available from: https://hbr.org/2017/03/the-blockchain-will-do-to-banks-and-law-firms-what-the-internet-did-to-media
-    \item U.S.
-Department of Education, Office of Educational Technology (2012).
-\textit{Enhancing teaching and learning through educational data mining and learning analytics: An issue brief}.
-U.S.
-Department of education: Washington, D.C.
-    \item EU General Data Protection Regulation (GDPR): Regulation (EU) 2016/679 of the European Parliament and of the Council of 27 April 2016 on the protection of natural persons with regard to the processing of personal data and on the free movement of such data, and repealing Directive 95/46/EC (General Data Protection Regulation), OJ 2016 L 119/1.
-    \item Kelly, M. (2019). New privacy bill would give parents an ‘Eraser Button’ and ban ads targeting children. \textit{The Verge}, March 12.
+It lives all around us: Aspects of data literacy in teen’s lives.
+\textit{Proceedings of the Association for Information Science and Technology, 54(1)}, 27–35.
+DOI:10.1002/pra2.2017.14505401004
+    \label{sec:26}
+    \item Friedman, B., and Nissenbaum, H. (1996). Bias in computer systems. \textit{ACM Transactions on Information Systems (TOIS), 14(3)}, 330-347.
+    \label{sec:27}
+    \item Borgman, C. L. (2015). \textit{Big data, little data, no data: Scholarship in the networked world.} MIT Press.
+    \label{sec:28}
+    \item Kumar, P., Naik, S., Devkar, U., Chetty, M., Tamara, C., and Vitak, J.
+(2017).
+'No telling passcodes out because they’re private’: Understanding children’s mental models of privacy and security online.
+\textit{Proceedings of the ACM on Human-Computer Interaction, 1(CSCW)}, 64.
+    \label{sec:29}
+    \item Azaria, A., Ekblaw, A., Vieira, T., & Lippman, A.
+(2016, August).
+Medrec: Using blockchain for medical data access and permission management.
+\textit{In 2016 2nd International Conference on Open and Big Data (OBD)} (pp.
+25-30).
+IEEE.
+    \label{sec:30}
+    \item Zeide, E. (2014). The proverbial permanent record.
 

### 2019-06-07 paper.tex
add final draft without formatting stuff

@@
 
-\title{Kratos: A solution for data privacy, literacy and student agency in a data-driven educational ecosystem}
+\title{Kratos: A solution for data privacy, literacy and student agency in a data-driven school ecosystem}
 \author{Dr.\ Velislava Hillman \quad Varunram Ganesh}
@@
 First, schools globally are equivocal about data generated by and about students as a result of the digitization of instruction, learning, and assessment.
-They lack necessary frameworks for data literacy, data interoperability, and optimization while maintaining privacy and control.
+They lack necessary frameworks for data literacy, data interoperability, and optimization, while maintaining privacy and control.
 Second, the scale, source, and nature of school data makes its interoperability impractical, resulting in an inability to assess the true impact of educational technologies on instruction and learning.
 Third, while data helps teachers improve pedagogical practices, an increasingly data-driven decision-making process suggests that student dimensions of learning and equitable participation in curriculum design becomes secondary.
-Finding a balance between data-driven decision making and student voice is critical for an efficient educational ecosystem.
-In this paper, we introduce Kratos: an immutable decentralised data management system that provides data privacy and applied data literacy while empowering students with a user interface for data governance and active participation in the educational ecosystem.
+Finding a balance between data-driven decision making and student voice is critical for an efficient school ecosystem.
+In this paper, we introduce Kratos: an immutable decentralised data management system that provides data privacy and applied data literacy while empowering students with a user interface for data governance and active participation in the school ecosystem.
 Using the advantages of blockchain technologies, Kratos enables easy authentication and access to data.
 The objective of Kratos is thus to equip schools and students with the ability to access, manage and control their data and to understand how, why, and by whom data is accessed without compromising student agency and privacy.
-This paper describes proof of concept for Kratos and its benefits to the educational ecosystem.
+This paper describes proof of concept for Kratos and its benefits to the school ecosystem.
 \bigbreak
-
+\newpage
 \section{Introduction}
 
-To improve work, school practitioners at local, district, state, and federal levels need data interoperability and educational information from data mining [Gates Foundation, 2015; US Dept of Education, 2012; ACT Policy Report, 2015].
-This need originally emanates from the United States federal law which mandates schools at district, state, and federal level to collect student data to serve as 'accountability' metric for assessing school effectiveness [Elementary and Secondary Education Act, 2001; 2010].
+To improve work, school practitioners need data interoperability and educational information from data mining (Bienkowski, Feng, and Means, 2012; Snyder et al., 2016).
+This need emanates from the requirement to measure school effectiveness by combining student achievement, data, and accountability (Militello et al., 2013).
 \bigbreak
 Educational data is important not only for evaluating school quality.
-It can also contribute to research in learning and advance theory [Baker, 2016].
-It can help guide intervention and lead to higher school attainment [Arnold and Pistill, 2012], improved pedagogy [Baker, 2014], and better parental involvement in school matters [Hawn, 2015].
-A common goal of educational data is to equip educators with key information that they can act upon and use to the benefit of learners [Baker, 2014].
-Educational data can enable an applied data literacy.
-Furthermore, access to school records provides an opportunity for student agency by enabling learning and participation.
+It can also contribute to research in learning and advance theory (Baker and Siemens, 2014).
+It can help guide intervention and lead to higher school attainment (Arnold and Pistilli, 2012), improved pedagogy (Baker and Siemens, 2014), and better parental involvement in school matters (Baker, 2016).
+A common goal of educational data is to equip educators with key information that they can act upon and use to the benefit of learners (Baker and Siemens, 2014).
+Educational data can enable an applied data literacy and train students to understand the benefits and limitations from data that is generated by and about them.
+Furthermore, access to educational data provides an opportunity for student agency by enabling interaction and participation.
 \bigbreak
-Increasingly digitized learning environments enable constant data collection and algorithm-based assessment which change core school functions of teaching, assessment, and accreditation [Zeide, 2017].
+Increasingly digitized school environments enable constant data collection, which change core school functions of teaching, assessment, and accreditation (Zeide, 2017).
 Digital systems for teaching and assessment drive data-based decision-making and enable micromanagement of students that can further pose restrictions on teacher autonomy and student (and parent) participation and willingness to challenge education decision-making.
-Schools deploy management systems for sorting and storing data and while some data is stored by schools at the district level, other data is managed by third-party vendors [Zhang, Li and Hao, 2015] who may consider this data proprietary and thus inaccessible.
-This puts barriers to effective data sharing and comprehensive understanding of school data.
+Schools deploy management systems for sorting and storing data and while some data is stored by schools at district level, other data is managed by third-party vendors who may consider this data proprietary and thus inaccessible by school practitioners, less so by students.
+This puts barriers to effective data sharing and comprehensive understanding and utilization of school data.
 \bigbreak
 In this work we propose the development of a structure for student data management, privacy, accountability, and auditability.
-We build on top of existing data standards and construct a common data schema for disparate data across different systems.
+We build on top of existing data standards and construct a common data schema for otherwise disparate data across different systems.
 Our system organises these references onto an integral structure of student data and complements the existing educational data standards and interoperability in three distinctive ways.
@@
 We design data analytic models to integrate with existing school systems and data standards.
-Lastly, we build a simple user interface that gives students, parents and schools access to disparate data, enabling control over what portion of the data can be shared.
+Lastly, we build a simple user interface that gives students, parents, and schools access to otherwise scattered and disparate data, enabling control over what portion of the data can be shared.
 Additionally, we propose an application with social functionality for student feedback.
 \bigbreak
-We present Kratos not as an independent solution to lack of data interoperability, accountability, and privacy but rather as a system that complements existing efforts and solutions developed by the various stakeholders in the educational ecosystem (Data Quality Standards, 2018; Project Unicorn; Bill and Melinda Gates Foundation, 2015).
+We propose this structure for K12 schools.
+However, the aim is that this initial work will pave the way for setting up a unique decentralised data management system that provides students with full control and visibility of data that is generated by or about them throughout their educational life.
+We acknowledge the legal, structural, technical, and organisational complexities involving school data of minors and therefore do not delve into these particular subjects.
+We focus on conceptualising how a safe space for student data can be created.
+We thus propose a tool for K12 students, which enables integrity and auditability of data that is collected about them.
+At the same time we project our initiative to develop into a platform, which will give students, once independent, full control over their data.
+\bigbreak
+We present Kratos not as an independent solution to lack of data interoperability, accountability, and privacy but rather as a system that complements existing efforts and solutions developed by various stakeholders in the school ecosystem (for example, Data Quality Campaign, 2018; Student Data Privacy Consortium, 2018).
 
 \section{Existing problems}
-To contextualise the complexity related to school data - the opportunities and challenges schools face as a result of the growing digitisation of operation and academic processes - Kratos has partnered with  Cambridge Public School (CPS) district that administer public elementary and high schools in Cambridge, Massachusetts, the Access for Learning (SIF), and the Student Data Privacy Consortium.
-With the help of CPS' Information and Communication Technology Services Chief Information Officer and Database Administrator, we explored data interoperability issues with them along with feedback on potential mitigation schemes:
+To contextualise the complexity related to school data - the opportunities and challenges schools face as a result of the growing digitisation of operation and academic processes - we partnered with Cambridge Public School district (CPS) that administer public elementary and high schools in Cambridge, Massachusetts, the Access for Learning (SIF), and the Student Data Privacy Consortium (SDPC), collaborative of schools, districts, policy makers, and other stakeholders addressing data privacy concerns (SDPC, 2018).
+With the help of CPS's Information and Communication Technology Services Chief Information Officer and Database Administrator, we explored data interoperability issues along with feedback on potential mitigation schemes:
 
@@
   \item \textbf{Data access} - CPS is in agreement with over 100 vendors providing education technologies.
-A majority of these provide no direct and comprehensive access to data generated about and by students who use of their products.
-Where available, data access is  provided in the form of reports - a summary of information which the district database administrator can request and download.
-On other occasions vendors supply data directly to teachers as well as the students in the form of digital dashboards.
-Where teachers obtain data it requires further work to convert it into meaningful information upon which the teacher can adjust and plan instruction (Data Quality Campaign, 2018).
+Many of them provide no direct and comprehensive access to data generated about and by students who use their products and services.
+Where available, data access is  provided in the form of reports - a summary of information, which the district database administrator can request and download.
+On other occasions, vendors supply data directly to teachers and in some cases to students in the form of digital dashboards.
+Where teachers obtain data, it requires further work to convert it into meaningful information upon which the teacher can adjust and plan instruction (Data Quality Campaign, 2018).
 
-  \item \textbf{Lack of data standard compliance} - School data frameworks vary across districts and states.
-Additionally, vendors providing education technologies use different data formats, schema and elements to organise and store student data as no single standard can be enforced upon them to comply with.
-Thus, disparate data scattered across different systems with varying degrees of accessibility and usability poses challenges to educators and other stakeholders in the education sector in assessing the impact of education technologies on the learning process.
-Moreover, the lack of data standard compliance further limits the potential for imposing privacy law as is the case with the General Data Protection Regulation (GDPR, 2016), which provides sound privacy practices that are, on the one hand, enabled by data interoperability, and on the other, imposing vendors to maintain data transparency.
+  \item \textbf{Lack of data standard compliance} - School data frameworks vary across districts and states (CEDS, no date).
+Additionally, vendors providing education technologies use different data formats, schema, and elements to organise and store student data as no single standard can be enforced upon them to comply with.
+Thus, disparate data, scattered across different systems with varying degrees of accessibility and usability poses challenges to educators and other stakeholders in the education sector in assessing the impact of education technologies on the learning process and on identifying best practices.
+The lack of data standard compliance impedes upon data transparency and prevents data interoperability, two conditions critical to ensuring data privacy.
 
   \item \textbf{Lack of transparency} - The scale, complexity, and number of providers pose challenges for schools to have a comprehensive list of the data that is generated about students.
-Lack of data interoperability standards which would simply be the process of compiling, organising and documenting what information is generated about students makes data auditability difficult to achieve, less so to be useful to the instruction and learning processes.
+Lack of data interoperability standards which is the process of compiling, organising, and documenting what information is generated about students makes data auditability difficult to achieve, less so to be useful to the instruction and learning processes.
+While some stakeholders in the education sector make tremendous efforts in ensuring data transparency in order to achieve effective privacy management (SDPC, 2018), issues persist (Kelly et al., 2018).
 
-  \item \textbf{Security} - Due to lack of transparency with regards to how various education technology vendors organise and structure their data, it becomes difficult to understand the security they ensure for the data their products and services generate.
-A recent case  presents the fears that exist with regards to school data (Cameron, 2017) - Schoolzilla, a K-12 education data service provider, detected a data breach which exposed the personal data of 1.3 million K-12 students.
-Some argue (Hand, 2018) that currently available cybersecurity protocols are inefficient to secure student data while at the same time allowing data analytics.
-Having inadequate access to available data further limits potentials to fully benefit from it.
+  \item \textbf{Security} - Due to general lack of transparency with regards to how various education technology vendors organise and structure their data (Ibid.), it becomes difficult to understand the security they ensure for the data their products and services generate.
+Having inadequate access to available data limits potentials to fully benefit from it.
 
@@
 
-We thus infer that schools are finding it increasingly difficult to obtain data about their own students in a comprehensive and timely manner and are thus becoming isolated due to a lack of knowledge on how vendors are processing the data that is generated as a result of their products and services used in class.
+We thus infer that schools generally experience numerous challenges that preclude them from obtaining student data in a comprehensive and timely manner, which further prevents students from understanding, accessing, interacting with, or having control (now or in the future) of their data.
 
@@
 
-Interoperability challenges between vendors and schools at local, district, state, or federal level pose barriers to cohesive data management and sharing while lacking basic technological privacy infrastructure and accountability (Zeide, 2014; Common Sense Media Research, 2018).
+Interoperability challenges between vendors and schools at local or district level pose barriers to cohesive data management and sharing while lacking basic technological privacy infrastructure and accountability (Zeide, 2014; Kelly et al., 2018).
 While digitizing and collecting student data is not new (Fitzgerald, 2014), technological developments like cloud computing and the Internet of Things (IoT) amplify concerns about data transfer, storage, use, and analysis (Sultan, 2010).
@@
 Concerns arise about over-prevalence of fine-grained data collected about students (Zeide, 2017) and the risk of creating a 'permanent record' that can impede upon learners' futures (Cody, 2013).
-The use of online platforms and applications in schools, often provided by for-profit vendors many of whom have unclear policies about data privacy (Common Sense Education, 2018), generates a continuous stream of data about students' behavior and performance at an unprecedented scale (Zeide, 2017).
+The use of online platforms and applications in schools, often provided by for-profit vendors, many of whom have unclear policies about data privacy (Kelly, et al., 2018), generates a continuous stream of data about students' behavior and performance at an unprecedented scale (Zeide, 2017).
 Big data complicates traditional understanding of what constitutes sensitive information and what information serves an educational purpose.
-Such data however, continues to drive decision-making of practitioners and vendors, marginalizing student dimensions of learning and equitable participation in curriculum design.
+Such data however, continues to drive decision-making of educators and vendors, marginalizing student dimensions of learning and equitable participation in curriculum design.
 \bigbreak
-The present challenges of fragmented data access, the lack of student agency and auditability and the lack of a concrete framework for data accessibility demand a socio-technological solution which can ensure that educational data can be used to help improve school processes while not diminishing student privacy, agency and future opportunities.
+The present challenges of fragmented data access, the lack of student agency and auditability, and the lack of a concrete framework for data accessibility demand a techno-social solution, which can ensure that educational data can be used to help improve school processes while not diminishing student privacy, agency, and future opportunities.
 
 \section{Kratos}
-Kratos aims to define a set of guidelines and principles that schools should follow and provides a platform for both schools and students to see what kind of data is being collected by vendors with the help of an immutable log for auditing student data changes, access, and use.
-The goal for students would be to see how their data is being used and above the age of 18, gain access to this data.
-We define this as Data Ownership and Visibility - how students can own the data after the age of 18 and how students can see what their data is being used for.
-The goal for the school would be to make disparate data from different vendors compatible with each other and we define this problem as Data Interoperability.
+Kratos aims to define a set of guidelines and principles that schools should follow and provides a platform for both schools and students to see what kind of data is being collected by vendors with the help of an immutable log for auditing student data changes, access, and use (Figure 1).
+The goal for students would be to see what data is being collected about them, how their data is being used, and at age of majority, gain full control of this data.
+We define this as Data Ownership and Visibility - how students can own their data after age of majority and how students can see what their data is being used for.
+We aim to create a comprehensive reference for schools to otherwise disparate data stored by different vendors, make it compatible with each other and we define this problem as Data Interoperability.
+For the purpose of our case study we work with two types of data samples - a sample from one vendor supplying applications for instruction and assessment for K12 students and data sample using SIF standards from our partnering school district of Cambridge, Massachusetts.
 
@@
 \subsection{Data Interoperability}
-A traditional approach to interoperability would be to force a given set of standards on third party vendors, so that they reference them.
+A traditional approach to interoperability would be to force a given set of standards on third party vendors so that they reference them.
 However, work done by standardization bodies like SIF and CEDS has shown that this is at best partially effective since there is no financial or social incentive for vendors to migrate to the proposed set of standards.
-Kratos attempts to solve this issue by proposing a solution where different fields described by different vendors can be mapped to a single underlying scheme defined by Kratos and third party vendors would not have to change their established data organisation and structure.
+Kratos attempts to solve this issue by proposing a solution where different fields described by different vendors can be mapped to a single underlying scheme defined by Kratos so that third party vendors would not have to change their established data organisation and structure.
+Again, for the purpose of describing conceptual work we refer to only a small sample of disparate data that is stored at the school's and the vendor's end.
 
@@
 As an example, let us assume the underlying fields defined by Kratos are Grade\_A, Grade\_B and Grade\_C.
-If a specific vendor has a different name for the same field (example GradeA, GradeB and GradeC), Kratos would ingest the data from the vendor defined fields and convert it to the ones defined by Kratos.
+If a specific vendor has a different name for the same field (example GradeA, GradeB and GradeC), Kratos would ingest the data from the fields defined by the vendor and convert it to the ones defined by Kratos.
 This should be done across different formats since the reports database administrators receive right now are formatted as JSON, CSV or are in the form of Excel sheets.
@@
 \bigbreak
-Kratos proposes that this be done in the form of a templated script that can be written for every vendor.
+We propose that this be done in the form of a templated script that can be written for every vendor.
 Building a common script for all the vendors poses a challenge since the number of vendors and the different fields and standards they follow are ever changing.
@@
 The number of fields however would change since SIF has over 700 different fields and it is not possible to accurately map all these fields.
-Kratos defines the idea of a "bucket" - a collection of data fields collated together as a single entity to make it easier for students and administrators to monitor them.
-Common examples of buckets would be PII (Personally Identifiable Information), Grade Reports (containing grade reports for each subject) and Attendance Records (recording the class wise attendance of the student).
-To formalise the type of data buckets it is imperative to develop a data taxonomy that comprises the existing school data standards and data elements according to which different education technology providers collect and organise their data.
+For Kratos we define the idea of a "bucket" - a collection of data fields collated together as a single entity to make it easier for students and administrators to monitor them.
+Common examples of buckets would be PII (Personally Identifiable Information), Grade Reports (containing grade reports and tests per subject) and Attendance Records (recording the class wise attendance of the student).
+Other buckets may include Metadata (recording student use and interaction with an application) and Conduct (including discipline, accidents, and other data relating to student behaviour).
+To formalise the type of data buckets, it is imperative to develop a data taxonomy - a comprehensive vocabulary of data type, purpose of use, and impact - comprising data collected by schools and education technology providers.
 
@@
 \bigbreak
-The buckets defined by Kratos can be used in multiple ways: students can have a better understanding of how their data is being used, parents can know what kind of data is being collected about their children, and administrators can ensure that sensitive information is not being shared with third-party vendors.
-After legal age of consent, this control would be given to the hands of students and they can decide if they wish to continue sharing data with these parties.
+The buckets defined by Kratos provide reference to corresponding and otherwise disparate data (including data residing with vendors).
+The data relating to each bucket (each of which further breaks down into concrete elements) can be used in multiple ways: students can have a better understanding of what data is generated about them, what for, and by whom their data is being used; parents can know what kind of data is being collected about their children; and school administrators can benefit from a comprehensive view and access to an ever growing trail of interoperable data.
 
@@
 In conventional systems, data control or 'ownership' can be proved with the help of an access token but this provides no guarantee on when the owner came into possession of the data.
-In order to attest 'ownership' at a specific point in time, we need time-stamping services like those described in \hyperref[sec:1]{[1]}, \hyperref[sec:2]{[2]} and \hyperref[sec:3]{[3]}.
+In order to attest 'ownership' at a specific point in time, we need time-stamping services like those described by Todd (2016), Szalachowski (2018), and Massias et al.(1999).
 A time-stamping service requires something to be committed along with the time-stamp and Kratos envisions this to be a cryptographic hash which also acts as an access token which vendors can use to access any student-generated data.
 \bigbreak
-The aim of using a cryptographic hash like SHA-3 \hyperref[sec:6]{[6]} is to ensure a uniquely random reference to the data to avoid out-of-channel data leaks.
+The aim of using a cryptographic hash like SHA-3 (Bertoni et al., 2012) is to ensure a uniquely random reference to the data to avoid out-of-channel data leaks.
 Kratos suggests using an element of randomness like a salt to generate the hash in order to have the ability to revoke tokens by regenerating randomness.
 Kratos enforces that all schools encrypt their data before creating access tokens to mitigate the risk of loss/theft of data.
-Past studies like \hyperref[sec:4]{[4]} and \hyperref[sec:5]{[5]} show that firms are willing to circumnavigate laws to collect data and encrypting data by design ensures that no third party can have access without being granted so explicitly.
+Past research (Zetter, 2016) shows that firms are willing to circumnavigate laws to collect data and encrypting data by design ensures that no third party can have access without being granted so explicitly.
 \bigbreak
@@
 In addition to this, Kratos suggests that schools store their encrypted data in a distributed file storage system like IPFS to ensure data redundancy and availability in case of a setback.
-Storing data on IPFS also makes it easier to create timestamps since it is sufficient to reference the IPFS pointer instead of potentially hashing the whole data.
+Storing data on IPFS also makes it easier to create time-stamps since it is sufficient to reference the IPFS pointer instead of potentially hashing the whole data.
 \bigbreak
-In the event a user wants to revoke access to a particular vendor, he could do so by changing the encryption key or by changing the randomness used to generate the access token.
-We suggest users do not regenerate encryption keys but Kratos will provide users an option to choose between the two.
+In the event that a user wants to revoke access to a particular vendor, he could do so by changing the encryption key or by changing the randomness used to generate the access token.
+We suggest that users do not regenerate encryption keys but Kratos will provide users an option to choose between the two.
+As we describe school data of K12 students in principle it is the school or school district administrator who controls such access.
+The difference is that the system will further establish accountability and auditability to students with the capacity to gain further control over their data when they become legally independent.
 \bigbreak
-Since schools have their own sets of policies, Kratos does not strictly enforce a set of practices for users to follow.
+Furthermore, since schools have their own sets of policies, Kratos does not strictly enforce a set of practices for users (at school or district level) to follow.
 This ensures that adoption of Kratos is not constrained by a certain set of rules.
@@
 \bigbreak
-Kratos also does not enforce how commitments need to be generated and stored, leaving it as an option for schools to provide their feedback on.
-Solutions provided by Kratos would include a centralized time-stamping server, a permissioned blockchain with the different schools as the nodes and simple time-stamping commitments to an existing blockchain.
+Kratos also does not enforce how commitments need to be generated and stored, leaving it as an option for schools to provide their feedback.
+The system we propose would include a centralized time-stamping server, a permissioned blockchain with the different schools as the nodes, and simple time-stamping commitments to an existing blockchain.
 All three have their benefits and constraints and Kratos would enable schools to customize this to their requirements.
@@
 \bigbreak
-While summative and cumulative assessments provide "empirically observed specific dimensions" about student academic performance, student agency and active participation is equally required in order for policy and education to design "strategies for change".
-Growing use of education technologies enables data-driven decision-making [Gibson et al., 2015].
-Technology-mediated instruction and assessment tools with learning analytics functionality track and diagnose student progress.
-Most educational technologies can interpret and equip educators with information via digital dashboards and 'skill meters'[Baker, 2016; New, 2016].
-Fine-grained and continuously accumulated data about student behavior and performance surpasses traditional notions of assessment [Zeide, 2017] posing limitations over student dimensions of learning and equitable participation in the curriculum design.
+While summative and formative assessments provide "empirically observed specific dimensions" about student academic performance, student agency and active participation is equally required in order for policy and education to design "strategies for change".
+Growing use of education technologies enables data-driven decision-making (Gibson et al., 2015).
+Technology-mediated instruction and assessment tools with learning analytics functionality track and diagnose student progress (Tempelaar et al., 2013).
+Most educational technologies can interpret and equip educators with information via digital dashboards and 'skill meters'(Baker, 2016).
+Fine-grained and continuously accumulated data about student behavior and performance surpasses traditional notions of assessment (Zeide, 2017) posing limitations over student dimensions of learning and equitable participation in the curriculum design.
 \bigbreak
-Our prototype provides a graphical interface for student involvement in and accessibility to school data.
-While students and equally their parents can become acquainted with any changes and meanings of school data, students can also participate with personal feedback to the learning process.
+Our prototype provides a graphical interface for student involvement in (Figures 3-8) and accessibility to school data.
+While students and equally their parents can become acquainted with any changes and meanings of school data, students can also participate with personal feedback in the learning process (Figure 8).
 Student participation with personal perspectives and reflection are integral to the learning process (Ackermann, 1996).
@@
 Both school climate surveys and the proposed student feedback application provide flexible selection of questionnaires and measurements with a common goal to improve school climate.
-However, the proposed application enables not only feedback from older students (Ibid.) but from all students, provided that the feedback addresses the goal to encourage perspective-taking and reflection that are deemed necessary to the learning process (Kegan, 1982).
+However, the proposed application enables not only feedback from older students (Ibid.) but from all students, provided that the feedback addresses the goal to encourage perspective-taking and reflection that support learning (Kegan, 1982).
 The application further enables student agency and control over the frequency, depth, and nature of the feedback provided that it directly reflects the learning experience.
@@
 \subsection{Data literacy}
- Data literacy is still a debatable construct (Bowler and Acker, 2017).
+ Data literacy is still a debatable construct (Bowler et al.,2017).
 Data means quantified information, which is "situated, taking its meaning from its context and the perspective of its beholder" (Borgman, 2015, p.
@@
 Making sense of data is relative.
-What data may mean to someone can mean nothing to others (Bowler and Acker, 2017).
+What data may mean to someone can mean nothing to others (Bowler et al.,, 2017).
 In increasingly data-driven systems it becomes imperative not only to understand data but to interact with it and participate in the decision-making processes that it increasingly begins to impact.
-While data provides insight about how systems operate, the inferences drawn from it is without context.
+While data provides insight about how systems operate, the inferences drawn from it can be without context.
 Fine-grained data collected over long periods of time can lead to negative impact on individual well-being and pose limitations over future opportunities (Altman et al., 2018; Friedman and Nissenbaum, 1996).
@@
  \bigbreak
- While some studies examine aspects of data literacy in young people's lives (Bowler and Acker, 2017; Selwyn and Pangrazio, 2018; Kumar et al., 2017), the focus remains on Internet and social media data.
+ While some studies examine aspects of data literacy in young people's lives (Bowler et al., 2017; Selwyn and Pangrazio, 2018; Kumar et al., 2017), the focus remains on data generated from the use of the Internet in general and social media specifically.
 Studies related to young people's perceptions, knowledge, and understanding of school data literacy are meagre.
  \bigbreak
- Existing literature demonstrates that young people have varying interprations and a general understanding of data.
-However, many still find difficulties to connect with data at a concrete level, with the notion of having a data dossier (Bowler and Acker, 2017, p.
+ Existing literature demonstrates that young people have varying interprations and a general understanding of data (Bowler et al., 2017).
+However, many still find difficulties to connect with data at a concrete level, with the notion of having a data dossier (Ibid., p.
 27).
@@
  Kratos thus proposes a comprehensive, yet simple access, knowledge, and interaction with school data.
-At conceptualisation stage we choose three core data elements to work with.
+At this conceptualisation stage, we choose three core data buckets to work with.
 These are assessment, including test scores, reports, and credentialing; conduct, including observation reports, behaviour, and attendance; and personally identifiable information, including personal identifiers, demographics, parent information, medical reports, and socio-economic information.
-We take these three data elements crucial to student privacy and control.
-In our conceptual work (Figures 3-5) we recognise that while schools collect data for each of these three elements, digital applications used in class also collect similar such data.
+We take these three data buckets to be crucial to student privacy and control.
+ \bigbreak
+ In our conceptual work (Figures 3-5) we recognise that while schools collect data for each of these three data buckets, digital applications used in class also collect similar such data.
 For example, as schools contain data about conduct, applications such as Class Dojo (Manolev et al., 2018) gather similar data.
-Teachers may further report behaviour using Swivl.
+Teachers may further report behaviour using Swivl, a video application that enables supervision and classroom observation.
 Currently, students do not have a comprehensive access to or knowledge of all data that exists about their conduct.
 Moreover, there is no awareness and knowledge of the potential impact of all data that exists about their conduct.
-We therefore create visibility to the disparate data that exists about student conduct, provide visibility and control over who accesses such data and for what purpose, and enable learning about data (figure 6), its meaning and purpose.
+We therefore create visibility to the disparate data that exists about student conduct, provide visibility, and learning about such data (Figure 6).
 
@@
     \includegraphics[height=9.5cm]{{"images/data-detail-1.png}}
-    \caption{User UI: Student view of data}
+    \caption{Data view}
     \label{fig:my_label}
@@
     \includegraphics[height=9.5cm]{{"images/data-detail-2.png}}
-    \caption{User UI: Student view of data and definition}
+    \caption{Data view with definition}
     \label{fig:my_label}
@@
     \includegraphics[height=9.5cm]{{"images/detail.png}}
-    \caption{Permit / Revoke Access}
+    \caption{Permit/revoke access}
     \label{Permit / Revoke Access}
@@
 \section{Discussion and future work}
-In this paper we propose Kratos, a decentralised data management system which enables data interoperability, applied data literacy, and student participation in the curriculum design.
-This proposal envisions an ideal techno-social solution that empowers students by providing more control over their school data through data transparency, accountability and auditability.
-At the same time, continuous legal (GDPR, 2016; Colorado Student Data Transparency and Security Act, 2016; Kelly, 2019) and policy (Student Data Privacy Consortium, 2018) efforts are being made that will further help establish Kratos as a tool integral to the school ecosystem.
+In this paper we propose Kratos, a decentralised data management system, which enables data interoperability, applied data literacy, and student participation.
+This proposal envisions a techno-social solution that empowers students by providing control over their school data through data transparency and auditability.
+While legal (Colorado Student Data Transparency and Security Act, 2016; Kelly, 2019) and policy (e.g.
+SDPC, 2018) efforts are being made to ensure data privacy and security, much more needs to be done (Bulger et al., 2019), which encourages us to continue work on the Kratos project as a technological layer of protection and data privacy.
 \bigbreak
-Unlike other sectors such as retail (Lu and Xu, 2017), finance (Ito et al., 2017), and medicine (Acbo et al., 2019), education is still not fully leveraging the power of global digital transformation.
-In medicine, for instance, data interoperability is important in order for various stakeholders to access comprehensive patient records and various cases across Europe (Demertzis et al., 2018) and the United States (Azaria et al., 2016) decentralised architectures have been developed to enable data sharing, interoperability, and access while prioritising patient agency, privacy, and security.
+Unlike other sectors such as retail (Lu and Xu, 2017), finance (Ito et al., 2017), and medicine (Agbo et al., 2019), education is still not fully leveraging the power of global digital transformation.
+In medicine, for instance, data interoperability is important in order for various stakeholders to access comprehensive patient records.
+Across Europe (Demertzis et al., 2018) and the United States (Azaria et al., 2016), decentralised architectures have been developed to enable data sharing, interoperability, and access while prioritising patient agency, privacy, and security.
 Similarly, a more comprehensive data management for interoperability is highly needed in the education sector without compromising student privacy and control over their data.
 \bigbreak
-On the other hand, while data analytics provides insight about school processes, not everything that is measured is meaningful.
-Students’ perspectives and reflection should equally count next to data-driven decision-making.
-Moreover, the sophistication of data analytics and long-term data collection further decrease an individual's privacy.
-The risk of being publicly exposed can lead to being embarrassed or discriminated (Altman et al.
+While our conceptual work focuses on K12 education we understand the complexities surrounding the sheer volume of data that is generated across many different systems, the differing interests of stakeholders in the education sector, and the hurdles involving access to data that many vendors consider proprietary.
+\bigbreak
+We propose Kratos as a decentralised system that enables data integrity and visibility of the huge data trail of students who are still dependent on adults to decide for them.
+However, such a system can give them an immutable log of their data as an additional layer of security now and at any point in time beyond their secondary and higher education.
+\bigbreak
+Comprehensive access to data opens the possibilities for better analysis of processes and systems (Baker and Siemens, 2014).
+However, while data analytics gives some insight, not everything that is measured is valuable and not everything that is of value can be measured.
+Students’ perspectives should go on a par with data-driven decision-making.
+Moreover, the sophistication of data analytics and long-term data collection can potentially decrease individual privacy (Dwork and Rothblum, 2016).
+The risk of being publicly exposed can lead to embarrassment or discrimination (Altman et al.
 2018).
-Constantly being aware of such risks can diminish one’s sense of freedom for self-expression and in the case with children and young people - the freedom to try out new things and make mistakes.
+Constantly being aware of such risks can diminish one’s sense of freedom for self-expression and in the case with children and young people - the freedom to try out new things and make mistakes, things that are integral to learning.
 Therefore, a decentralised data management system that prioritises student privacy and control over data, becomes crucial to the development of a safe and free learning environment.
@@
 First, we plan to develop a comprehensive taxonomy of school data.
-This will serve to develop a comprehensive data vocabulary for applied data literacy and identify data elements.
+This will serve to develop a comprehensive data vocabulary for applied data literacy and identify data elements - the trail and origin, and its purpose of use.
 The second step involves prototyping and formalising platform sections and functionality.
-And the third involves carrying out user studies to get student feedback and the fine tuning of the platform's user interface.
+And the third involves carrying out user studies to get student feedback and fine-tuning the platform's user interface.
+All future work is part of the greater effort to provide data privacy protection of students.
 
 \section{Acknowledgements}
-We thank Cambridge Public School district, Student Data Privacy Consortium, Access for Learning (SIF) for their support.
+We thank Cambridge Public School district, Student Data Privacy Consortium, and Access for Learning (SIF) for their support, constant, and timely feedback.
 We would also like to thank Rayner Ng Jing Kai and Geoffrey Martin from Yale-NUS College for their help in sketching mock-up designs and giving feedback on data interoperability.
+Lastly we greatly appreciate the invaluable feedback on decentralised ledger technologies from Mario Galea at Random Consulting.
 
@@
     \label{sec:1}
-    \item Todd, P. OpenTimestamps: Scalable, Trustless, Distributed Timestamping with Bitcoin (2016). URL \url{https://petertodd.org/2016/opentimestamps-announcement}
+    \item Todd, P. OpenTimestamps: Scalable, Trustless, Distributed Timestamping with Bitcoin (2016). URL \url{https://petertodd.org/2016/opentimestamps-announcemen}
     \label{sec:2}
@@
     \label{sec:5}
-    \item Dissent, Back-To-School Revolt in Springfield?
-Employees balk over using Google Drive as evidence of massive privacy breach mounts (2018).
-URL \url{https://www.pogowasright.org/back-to-school-revolt-in-springfield-employees-balk-over-using-google-drive-as-evidence-of-massive-privacy-breach-mounts/}
+    \item Singer, N. (2017). How Google took over the classroom. Available from: https://www.nytimes.com/2017/05/13/technology/google-education-chromebooks-schools.html
     \label{sec:6}
@@
     \label{sec:9}
-    Zeide, E. (2016). 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?.
+    Zeide, E. (2017). 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?.
     \label{sec:10}
@@
 C., Webb, M.
-E., & Ifenthaler, D.
+E., and Ifenthaler, D.
 (2019).
 Measurement Challenges of Interactive Educational Assessment.
-In \textit{Learning Technologies for Transforming Large-Scale Teaching, Learning, and Assessment} (pp.
-19-33).
+In \textit{Learning Technologies for Transforming Large-Scale Teaching, Learning, and Assessment}, 19-33.
 Springer, Cham.
@@
 C., Mahmoud, Q.
-H., & Eklund, J.
+H., and Eklund, J.
 M.
@@
     \label{sec:18}
-    \item Lu, Q., & Xu, X. (2017). Adaptable blockchain-based systems: a case study for product traceability. \textit{IEEE Software,} 34(6), 21-27.
+    \item Lu, Q., and Xu, X. (2017). Adaptable blockchain-based systems: a case study for product traceability. \textit{IEEE Software,} 34(6), 21-27.
     \label{sec:19}
-    \item Ito, J., Narula, N., & Ali, R. (2017). The blockchain will do to the financial system what the internet did to media. \textit{Harvard Business Review,} 9(March).
+    \item Ito, J., Narula, N., and Ali, R. (2017). The blockchain will do to the financial system what the internet did to media. \textit{Harvard Business Review,} 9(March).
     \label{sec:20}
-    \item Bienkowski, M., Feng, M., & Means, B.
+    \item Bienkowski, M., Feng, M., and Means, B.
 (2012).
@@
 P.
+R.
 (2016).
@@
     \item Kelly, M.
-New privacy bill would give parents an ‘Eraser Button’ and ban ads targeting children (2019).
+(2019).
+New privacy bill would give parents an ‘Eraser Button’ and ban ads targeting children.
 URL \url{https://www.theverge.com/2019/3/12/18261181/eraser-button-bill-children-privacy-coppa-hawley-markey}
@@
     \label{sec:24}
-    \item Altman, M., Wood, A. B., O'Brien, D., & Gasser, U. (2018). Practical approaches to big data privacy over time.
+    \item Altman, M., Wood, A. B., O'Brien, D., and Gasser, U. (2018). Practical approaches to big data privacy over time.
     \label{sec:25}
@@
     \label{sec:29}
-    \item Azaria, A., Ekblaw, A., Vieira, T., & Lippman, A.
+    \item Azaria, A., Ekblaw, A., Vieira, T., and Lippman, A.
 (2016, August).
-Medrec: Using blockchain for medical data access and permission management.
-\textit{In 2016 2nd International Conference on Open and Big Data (OBD)} (pp.
-25-30).
+MedRec: Using blockchain for medical data access and permission management.
+\textit{In 2016 2nd International Conference on Open and Big Data (OBD)},  25-30.
 IEEE.
@@
     \item Zeide, E. (2014). The proverbial permanent record.
+    \label{sec:31}
+    \item Common Education Data Standards - CEDS 101. Available from: https://ceds.ed.gov/pdf/ceds-101.pdf
+    \label{sec:32}
+    \item Tempelaar, D.
+T., Heck, A.
+J.
+P., Cuypers, H., van der Kooij, H., and van de Vrie, E.
+(2013).
+Formative assessment and learning analytics.
+in D.
+Suthers, K.
+Verbert, E.
+Duval, and X.
+Ochoa (Eds.), LAK 2013: Third International Conference on Learning Analytics and Knowledge: 8-12 April, 2013, Leuven, Belgium, 205-209.
+New York: Association for Computing Machinery.
+    \label{sec:33}
+    \item Bulger, M., Collins, S., Sallay, D., Vance, A.
+(2019).
+School safety and privacy: An animated introduction.
+\textit{FERPA, SHERPA: The Education Privacy Resource Centre}.
+Available from: https://ferpasherpa.org/schoolsafetyvideo/
+    \label{34}
+    \item Baker, R.
+and Siemens, G.
+(2014).
+Educational data mining and learning analytics.
+In Sawyer K.
+(Ed.), Cambridge handbook of the learning sciences (2nd ed.), 253–274.
+New York, NY: Cambridge University Press.
+    \label{35}
+    \item Snyder, T.D., de Brey, C., and Dillow, S.A.
+(2016).
+Digest of Education Statistics 2015 (NCES 2016-014).
+National Center for Education Statistics, Institute of Education Sciences, U.S.
+Department of Education: Washington, DC.
+    \label{36}
+    \item Militello, M., Bass, L., Jackson, T. K., and Wang, Y. (2013). How data are used and misused in schools: Perceptions from teachers and principals.\textit{Education Sciences, 3}, 98-120.
+    \label{37}
+    \item Arnold, K.
+E.
+and Pistilli, M.
+D.
+(2012, April).
+Course signals at Purdue: Using learning analytics to increase student success.
+Proceedings of the 2nd International Conference on Learning Analytics and Knowledge, 267–270.
+    \label{38}
+    \item Baker, R. S. (2016). Stupid tutoring systems, intelligent humans. \textit{International Journal of Artificial Intelligence in Education, 26}, 600-614.
+    \label{39}
+    \item Kelly, G., Graham, J., and Fitzgerald, B. (2018). \textit{State of edtech privacy report.} Common Sense Privacy Evaluation Initiative. San Francisco, CA: Common Sense. 
+    \label{40}
+    \item Data Quality Campaign.
+(2018).
+What parents and teachers think about education data.
+Available from: https://dataqualitycampaign.org/resource/what-parents-and-teachers-think-about-education-data/
+    \label{41}
+    \item Student Data Privacy Consortium.
+(2018).
+Student Data Privacy Consortium: Policy and procedures.
+SDPC.
+Available from: https://privacy.a4l.org/wp-content/uploads/2018/06/Student-Data-Privacy-Consortium-P-and-P-2018-Final.pdf
+    \label{42}
+    \item California Consumer Privacy Act of 2018 (CCPA), CAL. CIV. CODE § 1798.198(a) (2018).
+    \label{43}
+    \item Colorado Student Data Transparency and Security Act (HB 16-1423). (2016). Available from: https://www.cde.state.co.us/dataprivacyandsecurity/crs22-16-101
+    \label{44}
+    \item Agbo, C., Mahmoud, Q., and Eklund, J. (2019). Blockchain technology in healthcare: A systematic review. \textit{Healthcare, 7}, 56, 1-30.
+    \label{45}
+    \item Dwork, C., and Rothblum, G. (2016). Concentrated differential privacy. Working paper. Available from: https://arxiv.org/pdf/1603.01887.pdf
+    \label{46}
+    \item Sirota, D.
+(2013).
+Big data means kids' "permanent records" might never be erased.
+MOTHERBOARD, October 24.
+Available from: http://motherboard.vice.com/blog/permanent-records-are-hurting-kids.
+    \label{47}
+    \item Selwyn, N., and Pangrazio, L.
+(2018).
+Doing data differently?
+Developing personal data tactics and strategies amongst young mobile media users.
+\textit{ Big Data and Society}, January-June, 1-12.
+    \label{48}
+    \item Manolev, J. Sullivan A., and Slee, R. (2018). The datafication of discipline: Class Dojo, surveillance and a performative classroom culture. Learning, Media and Technology, 44(1), 36-51.
 

### 2019-06-07 paper.tex
add . and , formatting changes

--- 2019-06-05 paper.tex
@@
 
-\title{Kratos: blockchain solution for data privacy, literacy, and student agency in a data-driven educational ecosystem}
-\author{Dr.\ Velislava Hillman \quad Varunram Ganesh \quad Geoffrey Martin}
-\date{May 2019}
+\title{Kratos: A solution for data privacy,\ literacy and student agency in a data-driven school ecosystem}
+\author{Dr.\ Velislava Hillman \quad Varunram Ganesh}
+\date{June 2019}
 
@@
 \textit{Abstract}
-Growing digitization has made data ownership an important focus point for institutions and students.
-Broadly, there are three issues, which require urgent attention for optimization of data privacy, literacy, and utilization.
-First, schools globally are equivocal about data generated by and about students as a result of the digitization of instruction, learning, and assessment.
-They lack necessary frameworks for data literacy, data interoperability, and optimization while maintaining privacy and control.
-Second, the scale, source, and nature of school data makes its interoperability impractical, resulting in an inability to assess the true impact of educational technologies on instruction and learning.
-Third, while data helps teachers improve pedagogical practices, an increasingly data-driven decision-making process suggests that student dimensions of learning and equitable participation in curriculum design becomes secondary.
-Finding a balance between data-driven decision making and student voice is critical for an efficient educational ecosystem.
-In this paper, we introduce Kratos: an immutable decentralised data management system that provides data privacy and applied data literacy while empowering students with a user interface for data governance and active participation in the educational ecosystem.
-Using the advantages of blockchain technologies, Kratos enables easy authentication and access to data.
-The objective of Kratos is thus to equip schools and students with the ability to access, manage and control their data and to understand how, why, and by whom data is accessed without compromising student agency and privacy.
-This paper describes proof of concept for Kratos and its benefits to the educational ecosystem.
+Growing digitization has made data ownership an important focus point for institutions and students.\ Broadly,\ there are three issues,\ which require urgent attention for optimization of data privacy,\ literacy,\ and utilization.\ First,\ schools globally are equivocal about data generated by and about students as a result of the digitization of instruction,\ learning,\ and assessment.\ They lack necessary frameworks for data literacy,\ data interoperability,\ and optimization,\ while maintaining privacy and control.\ Second,\ the scale,\ source,\ and nature of school data makes its interoperability impractical,\ resulting in an inability to assess the true impact of educational technologies on instruction and learning.\ Third,\ while data helps teachers improve pedagogical practices,\ an increasingly data-driven decision-making process suggests that student dimensions of learning and equitable participation in curriculum design becomes secondary.\ Finding a balance between data-driven decision making and student voice is critical for an efficient school ecosystem.\ In this paper,\ we introduce Kratos: an immutable decentralised data management system that provides data privacy and applied data literacy while empowering students with a user interface for data governance and active participation in the school ecosystem.\ Using the advantages of blockchain technologies,\ Kratos enables easy authentication and access to data.\ The objective of Kratos is thus to equip schools and students with the ability to access,\ manage and control their data and to understand how,\ why,\ and by whom data is accessed without compromising student agency and privacy.\ This paper describes proof of concept for Kratos and its benefits to the school ecosystem.
 \bigbreak
-
+\newpage
 \section{Introduction}
 
-To improve work, school practitioners at local, district, state, and federal levels need data interoperability and educational information from data mining [Gates Foundation, 2015; US Dept of Education, 2012; ACT Policy Report, 2015].
-This need originally emanates from the United States federal law which mandates schools at district, state, and federal level to collect student data to serve as 'accountability' metric for assessing school effectiveness [Elementary and Secondary Education Act, 2001; 2010].
+To improve work,\ school practitioners need data interoperability and educational information from data mining (Bienkowski,\ Feng,\ and Means,\ 2012; Snyder et al.,\ 2016).\ This need emanates from the requirement to measure school effectiveness by combining student achievement,\ data,\ and accountability (Militello et al.,\ 2013).
 \bigbreak
-Educational data is important not only for evaluating school quality.
-It can also contribute to research in learning and advance theory [Baker, 2016].
-It can help guide intervention and lead to higher school attainment [Arnold and Pistill, 2012], improved pedagogy [Baker, 2014], and better parental involvement in school matters [Hawn, 2015].
-A common goal of educational data is to equip educators with key information that they can act upon and use to the benefit of learners [Baker, 2014].
-Educational data can enable an applied data literacy.
-Furthermore, access to school records provides an opportunity for student agency by enabling learning and participation.
+Educational data is important not only for evaluating school quality.\ It can also contribute to research in learning and advance theory (Baker and Siemens,\ 2014).\ It can help guide intervention and lead to higher school attainment (Arnold and Pistilli,\ 2012),\ improved pedagogy (Baker and Siemens,\ 2014),\ and better parental involvement in school matters (Baker,\ 2016).\ A common goal of educational data is to equip educators with key information that they can act upon and use to the benefit of learners (Baker and Siemens,\ 2014).\ Educational data can enable an applied data literacy and train students to understand the benefits and limitations from data that is generated by and about them.\ Furthermore,\ access to educational data provides an opportunity for student agency by enabling interaction and participation.
 \bigbreak
-Increasingly digitized learning environments enable constant data collection and algorithm-based assessment which change core school functions of teaching, assessment, and accreditation [Zeide, 2017].
-Digital systems for teaching and assessment drive data-based decision-making and enable micromanagement of students that can further pose restrictions on teacher autonomy and student (and parent) participation and willingness to challenge education decision-making.
-Schools deploy management systems for sorting and storing data and while some data is stored by schools at the district level, other data is managed by third-party vendors [Zhang, Li and Hao, 2015] who may consider this data proprietary and thus inaccessible.
-This puts barriers to effective data sharing and comprehensive understanding of school data.
+Increasingly digitized school environments enable constant data collection,\ which change core school functions of teaching,\ assessment,\ and accreditation (Zeide,\ 2017).\ Digital systems for teaching and assessment drive data-based decision-making and enable micromanagement of students that can further pose restrictions on teacher autonomy and student (and parent) participation and willingness to challenge education decision-making.\ Schools deploy management systems for sorting and storing data and while some data is stored by schools at district level,\ other data is managed by third-party vendors who may consider this data proprietary and thus inaccessible by school practitioners,\ less so by students.\ This puts barriers to effective data sharing and comprehensive understanding and utilization of school data.
 \bigbreak
-In this work we propose the development of a structure for student data management, privacy, accountability, and auditability.
-We build on top of existing data standards and construct a common data schema for disparate data across different systems.
-Our system organises these references onto an integral structure of student data and complements the existing educational data standards and interoperability in three distinctive ways.
-Through network permissioning and proofs of ownership on a distributed ledger, we enable data auditability and accountability.
-We design data analytic models to integrate with existing school systems and data standards.
-Lastly, we build a simple user interface that gives students, parents and schools access to disparate data, enabling control over what portion of the data can be shared.
-Additionally, we propose an application with social functionality for student feedback.
+In this work we propose the development of a structure for student data management,\ privacy,\ accountability,\ and auditability.\ We build on top of existing data standards and construct a common data schema for otherwise disparate data across different systems.\ Our system organises these references onto an integral structure of student data and complements the existing educational data standards and interoperability in three distinctive ways.\ Through network permissioning and proofs of ownership on a distributed ledger,\ we enable data auditability and accountability.\ We design data analytic models to integrate with existing school systems and data standards.\ Lastly,\ we build a simple user interface that gives students,\ parents,\ and schools access to otherwise scattered and disparate data,\ enabling control over what portion of the data can be shared.\ Additionally,\ we propose an application with social functionality for student feedback.\
 \bigbreak
-We present Kratos not as an independent solution to lack of data interoperability, accountability, and privacy but rather as a system that complements existing efforts and solutions developed by the various stakeholders in the educational ecosystem (Data Quality Standards, 2018; Project Unicorn; Bill and Melinda Gates Foundation, 2015).
+We propose this structure for K12 schools.\ However,\ the aim is that this initial work will pave the way for setting up a unique decentralised data management system that provides students with full control and visibility of data that is generated by or about them throughout their educational life.\ We acknowledge the legal,\ structural,\ technical,\ and organisational complexities involving school data of minors and therefore do not delve into these particular subjects.\ We focus on conceptualising how a safe space for student data can be created.\ We thus propose a tool for K12 students,\ which enables integrity and auditability of data that is collected about them.\ At the same time we project our initiative to develop into a platform,\ which will give students,\ once independent,\ full control over their data.\
+\bigbreak
+We present Kratos not as an independent solution to lack of data interoperability,\ accountability,\ and privacy but rather as a system that complements existing efforts and solutions developed by various stakeholders in the school ecosystem (for example,\ Data Quality Campaign,\ 2018; Student Data Privacy Consortium,\ 2018).
 
 \section{Existing problems}
-To contextualise the complexity related to school data - the opportunities and challenges schools face as a result of the growing digitisation of operation and academic processes - Kratos has partnered with  Cambridge Public School (CPS) district that administer public elementary and high schools in Cambridge, Massachusetts, the Access for Learning (SIF), and the Student Data Privacy Consortium.
-With the help of CPS' Information and Communication Technology Services Chief Information Officer and Database Administrator, we explored data interoperability issues with them along with feedback on potential mitigation schemes:
+To contextualise the complexity related to school data - the opportunities and challenges schools face as a result of the growing digitisation of operation and academic processes - we partnered with Cambridge Public School district (CPS) that administer public elementary and high schools in Cambridge,\ Massachusetts,\ the Access for Learning (SIF),\ and the Student Data Privacy Consortium (SDPC),\ collaborative of schools,\ districts,\ policy makers,\ and other stakeholders addressing data privacy concerns (SDPC,\ 2018).\ With the help of CPS's Information and Communication Technology Services Chief Information Officer and Database Administrator,\ we explored data interoperability issues along with feedback on potential mitigation schemes:
 
 \begin{itemize}
-  \item \textbf{Data access} - CPS is in agreement with over 100 vendors providing education technologies ([7]).
-A majority of these provide no direct and comprehensive access to data generated about and by students who use of their products.
-Where available, data access is  provided in the form of reports - a summary of information which the district database administrator can request and download.
-On other occasions vendors supply data directly to teachers as well as the students in the form of digital dashboards.
-Where teachers obtain data it requires further work to convert it into meaningful information upon which the teacher can adjust and plan instruction (Data Quality Campaign, 2018).
+  \item \textbf{Data access} - CPS is in agreement with over 100 vendors providing education technologies.\ Many of them provide no direct and comprehensive access to data generated about and by students who use their products and services.\ Where available,\ data access is  provided in the form of reports - a summary of information,\ which the district database administrator can request and download.\ On other occasions,\ vendors supply data directly to teachers and in some cases to students in the form of digital dashboards.\ Where teachers obtain data,\ it requires further work to convert it into meaningful information upon which the teacher can adjust and plan instruction (Data Quality Campaign,\ 2018).\
 
-  \item \textbf{Lack of data standard compliance} - School data frameworks vary across districts and states.
-Additionally, vendors providing education technologies use different data formats, schema and elements to organise and store student data as no single standard can be enforced upon them to comply with.
-Thus, disparate data scattered across different systems with varying degrees of accessibility and usability poses challenges to educators and other stakeholders in the education sector in assessing the impact of education technologies on the learning process.
-Moreover, the lack of data standard compliance further limits the potential for imposing privacy law as is the case with the General Data Protection Regulation (GDPR), which provides sound privacy practices that are, on the one hand, enabled by data interoperability, and on the other, imposing vendors to maintain data transparency.
+  \item \textbf{Lack of data standard compliance} - School data frameworks vary across districts and states (CEDS,\ no date).\ Additionally,\ vendors providing education technologies use different data formats,\ schema,\ and elements to organise and store student data as no single standard can be enforced upon them to comply with.\ Thus,\ disparate data,\ scattered across different systems with varying degrees of accessibility and usability poses challenges to educators and other stakeholders in the education sector in assessing the impact of education technologies on the learning process and on identifying best practices.\ The lack of data standard compliance impedes upon data transparency and prevents data interoperability,\ two conditions critical to ensuring data privacy.\
 
-  \item \textbf{Lack of transparency} - The scale, complexity, and number of providers pose challenges for schools to have a comprehensive list of the data that is generated about students.
-Lack of data interoperability standards which would simply be the process of compiling, organising and documenting what information is generated about students makes data auditability difficult to achieve, less so to be useful to the instruction and learning processes.
+  \item \textbf{Lack of transparency} - The scale,\ complexity,\ and number of providers pose challenges for schools to have a comprehensive list of the data that is generated about students.\ Lack of data interoperability standards which is the process of compiling,\ organising,\ and documenting what information is generated about students makes data auditability difficult to achieve,\ less so to be useful to the instruction and learning processes.\ While some stakeholders in the education sector make tremendous efforts in ensuring data transparency in order to achieve effective privacy management (SDPC,\ 2018),\ issues persist (Kelly et al.,\ 2018).
 
-  \item \textbf{Security} - Due to lack of transparency with regards to how various education technology vendors organise and structure their data, it becomes difficult to understand the security they ensure for the data their products and services generate.
-A recent case  presents the fears that exist with regards to school data (Cameron, 2017) - Schoolzilla, a K-12 education data service provider, detected a data breach which exposed the personal data of 1.3 million K-12 students.
-Some argue (Hand, 2018) that currently available cybersecurity protocols are inefficient to secure student data while at the same time allowing data analytics.
-Having inadequate access to available data further limits potentials to fully benefit from it.
+  \item \textbf{Security} - Due to general lack of transparency with regards to how various education technology vendors organise and structure their data (Ibid.),\ it becomes difficult to understand the security they ensure for the data their products and services generate.\ Having inadequate access to available data limits potentials to fully benefit from it.
 
@@
 
-We thus infer that schools are finding it increasingly difficult to obtain data about their own students and are becoming isolated due to a lack of knowledge on how vendors are processing this data.
+We thus infer that schools generally experience numerous challenges that preclude them from obtaining student data in a comprehensive and timely manner,\ which further prevents students from understanding,\ accessing,\ interacting with,\ or having control (now or in the future) of their data.\
 
@@
 
-Interoperability challenges between vendors and schools at local, district, state, or federal level pose barriers to cohesive data management and sharing while lacking basic technological privacy infrastructure and accountability [Zeide, 2014; Common Sense Media Research, 2018].
-While digitizing and collecting student data is not new [Fitzgerald, 2014], technological developments like cloud computing and the Internet of Things (IoT) amplify concerns about data transfer, storage, use, and analysis [Sultan, 2010].
+Interoperability challenges between vendors and schools at local or district level pose barriers to cohesive data management and sharing while lacking basic technological privacy infrastructure and accountability (Zeide,\ 2014; Kelly et al.,\ 2018).\ While digitizing and collecting student data is not new (Fitzgerald,\ 2014),\ technological developments like cloud computing and the Internet of Things (IoT) amplify concerns about data transfer,\ storage,\ use,\ and analysis (Sultan,\ 2010).
 \bigbreak
-Concerns arise about over-prevalence of fine-grained data collected about students [Zeide, 2017] and the risk of creating a 'permanent record' that can impede upon learners' futures [Cody, 2013].
-The use of online platforms and applications in schools, often provided by for-profit vendors many of whom have unclear policies about data privacy [Common Sense Education, 2018], generates a continuous stream of data about students' behavior and performance at an unprecedented scale [Zeide, 2017].
-Big data complicates traditional understanding of what constitutes sensitive information and what information serves an educational purpose.
-Such data however, continues to drive decision-making of practitioners and vendors, marginalizing student dimensions of learning and equitable participation in curriculum design.
+Concerns arise about over-prevalence of fine-grained data collected about students (Zeide,\ 2017) and the risk of creating a 'permanent record' that can impede upon learners' futures (Cody,\ 2013).\ The use of online platforms and applications in schools,\ often provided by for-profit vendors,\ many of whom have unclear policies about data privacy (Kelly,\ et al.,\ 2018),\ generates a continuous stream of data about students' behavior and performance at an unprecedented scale (Zeide,\ 2017).\ Big data complicates traditional understanding of what constitutes sensitive information and what information serves an educational purpose.\ Such data however,\ continues to drive decision-making of educators and vendors,\ marginalizing student dimensions of learning and equitable participation in curriculum design.
 \bigbreak
-The present challenges of fragmented data access, the lack of student agency and auditability and the lack of a concrete framework for data accessibility demand a socio-technological solution which can ensure that educational data can be used to help improve school processes while not diminishing student privacy, agency and future opportunities.
+The present challenges of fragmented data access,\ the lack of student agency and auditability,\ and the lack of a concrete framework for data accessibility demand a techno-social solution,\ which can ensure that educational data can be used to help improve school processes while not diminishing student privacy,\ agency,\ and future opportunities.
 
 \section{Kratos}
-Kratos aims to define a set of guidelines and principles that schools should follow and provides a platform for both schools and students to see what kind of data is being collected by vendors with the help of an immutable log for auditing student data changes, access, and use.
-The goal for students would be to see how their data is being used and above the age of 18, gain access to this data.
-We define this as Data Ownership and Visibility - how students can own the data after the age of 18 and how students can see what their data is being used for.
-The goal for the school would be to make disparate data from different vendors compatible with each other and we define this problem as Data Interoperability.
+Kratos aims to define a set of guidelines and principles that schools should follow and provides a platform for both schools and students to see what kind of data is being collected by vendors with the help of an immutable log for auditing student data changes,\ access,\ and use (Figure 1).\ The goal for students would be to see what data is being collected about them,\ how their data is being used,\ and at age of majority,\ gain full control of this data.\ We define this as Data Ownership and Visibility - how students can own their data after age of majority and how students can see what their data is being used for.\ We aim to create a comprehensive reference for schools to otherwise disparate data stored by different vendors,\ make it compatible with each other and we define this problem as Data Interoperability.\ For the purpose of our case study we work with two types of data samples - a sample from one vendor supplying applications for instruction and assessment for K12 students and data sample using SIF standards from our partnering school district of Cambridge,\ Massachusetts.
 
@@
 \subsection{Data Interoperability}
-A traditional approach to interoperability would be to force a given set of standards on third party vendors, so that they reference them.
-However, work done by standardization bodies like SIF and CEDS has shown that this is at best partially effective since there is no financial or social incentive for vendors to migrate to the proposed set of standards.
-Kratos attempts to solve this issue by proposing a solution where different fields described by different vendors can be mapped to a single underlying scheme defined by Kratos and third party vendors would not have to worry about using disparate fields.
+A traditional approach to interoperability would be to force a given set of standards on third party vendors so that they reference them.\ However,\ work done by standardization bodies like SIF and CEDS has shown that this is at best partially effective since there is no financial or social incentive for vendors to migrate to the proposed set of standards.\ Kratos attempts to solve this issue by proposing a solution where different fields described by different vendors can be mapped to a single underlying scheme defined by Kratos so that third party vendors would not have to change their established data organisation and structure.\ Again,\ for the purpose of describing conceptual work we refer to only a small sample of disparate data that is stored at the school's and the vendor's end.
 
@@
 \bigbreak
-As an example, let us assume the underlying fields defined by Kratos are Grade\_A, Grade\_B and Grade\_C.
-If a specific vendor has a different name for the same field (example GradeA, GradeB and GradeC), Kratos would ingest the data from the vendor defined fields and convert it to the ones defined by Kratos.
-This should be done across different formats since the reports database administrators receive right now are formatted as JSON, CSV or are in the form of Excel sheets.
+As an example,\ let us assume the underlying fields defined by Kratos are Grade\_A,\ Grade\_B and Grade\_C.\ If a specific vendor has a different name for the same field (example GradeA,\ GradeB and GradeC),\ Kratos would ingest the data from the fields defined by the vendor and convert it to the ones defined by Kratos.\ This should be done across different formats since the reports database administrators receive right now are formatted as JSON,\ CSV or are in the form of Excel sheets.
 
 \bigbreak
-Kratos proposes that this be done in the form of a templated script that can be written for every vendor.
-Building a common script for all the vendors poses a challenge since the number of vendors and the different fields and standards they follow are ever changing.
-This templated script would be triggered automatically by Kratos each time it receives an incoming report and the report itself will be parsed to understand which vendor it originates from.
+We propose that this be done in the form of a templated script that can be written for every vendor.\ Building a common script for all the vendors poses a challenge since the number of vendors and the different fields and standards they follow are ever changing.\ This templated script would be triggered automatically by Kratos each time it receives an incoming report and the report itself will be parsed to understand which vendor it originates from.
 
 \bigbreak
-The fields that Kratos defines will be adapted from the existing SIF standard followed by schools in Cambridge and would also have routine input from various experts on the subject.
-The number of fields however would change since SIF has over 700 different fields and it is not possible to accurately map all these fields.
-Kratos defines the idea of a "bucket" - a collection of data fields collated together as a single entity to make it easier for students and administrators to monitor them.
-Common examples of buckets would be PII (Personally Identifiable Information), Grade Reports (containing grade reports for each subject) and Attendance Records (recording the class wise attendance of the student).
-To formalise the type of data buckets it is imperative to develop a data taxonomy that comprises the existing school data standards and data elements according to which different education technology providers collect and organise their data.
+The fields that Kratos defines will be adapted from the existing SIF standard followed by schools in Cambridge and would also have routine input from various experts on the subject.\ The number of fields however would change since SIF has over 700 different fields and it is not possible to accurately map all these fields.\ For Kratos we define the idea of a "bucket" - a collection of data fields collated together as a single entity to make it easier for students and administrators to monitor them.\ Common examples of buckets would be PII (Personally Identifiable Information),\ Grade Reports (containing grade reports and tests per subject) and Attendance Records (recording the class wise attendance of the student).\ Other buckets may include Metadata (recording student use and interaction with an application) and Conduct (including discipline,\ accidents,\ and other data relating to student behaviour).\ To formalise the type of data buckets,\ it is imperative to develop a data taxonomy - a comprehensive vocabulary of data type,\ purpose of use,\ and impact - comprising data collected by schools and education technology providers.
 
 \bigbreak
-The models and scripts developed as part of enabling interoperability will be open source and subject to continuous updates.
-We would also be having a complete code audit before releasing the system in a production environment to ensure the model performs and behaves the way it was designed and intended to.
+The models and scripts developed as part of enabling interoperability will be open source and subject to continuous updates.\ We would also be having a complete code audit before releasing the system in a production environment to ensure the model performs and behaves the way it was designed and intended to.
 
 \bigbreak
-The buckets defined by Kratos can be used in multiple ways: students can have a better understanding of how their data is being used, parents can know what kind of data is being collected about their children, and administrators can ensure that sensitive information is not being shared with third-party vendors.
-After legal age of consent, this control would be given to the hands of students and they can decide if they wish to continue sharing data with these parties.
+The buckets defined by Kratos provide reference to corresponding and otherwise disparate data (including data residing with vendors).\ The data relating to each bucket (each of which further breaks down into concrete elements) can be used in multiple ways: students can have a better understanding of what data is generated about them,\ what for,\ and by whom their data is being used; parents can know what kind of data is being collected about their children; and school administrators can benefit from a comprehensive view and access to an ever growing trail of interoperable data.\
 
 \subsection{Data control}
-In conventional systems, data control or 'ownership' can be proved with the help of an access token but this provides no guarantee on when the owner came into possession of the data.
-In order to attest 'ownership' at a specific point in time, we need time-stamping services like those described in \hyperref[sec:1]{[1]}, \hyperref[sec:2]{[2]} and \hyperref[sec:3]{[3]}.
-A time-stamping service requires something to be committed along with the time-stamp and Kratos envisions this to be a cryptographic hash which also acts as an access token which vendors can use to access any student-generated data.
+In conventional systems,\ data control or 'ownership' can be proved with the help of an access token but this provides no guarantee on when the owner came into possession of the data.\ In order to attest 'ownership' at a specific point in time,\ we need time-stamping services like those described by Todd (2016),\ Szalachowski (2018),\ and Massias et al.(1999).\ A time-stamping service requires something to be committed along with the time-stamp and Kratos envisions this to be a cryptographic hash which also acts as an access token which vendors can use to access any student-generated data.
 \bigbreak
-The aim of using a cryptographic hash like SHA-3 \hyperref[sec:6]{[6]} is to ensure a uniquely random reference to the data to avoid out-of-channel data leaks.
-Kratos suggests using an element of randomness like a salt to generate the hash in order to have the ability to revoke tokens by regenerating randomness.
-Kratos enforces that all schools encrypt their data before creating access tokens to mitigate the risk of loss/theft of data.
-Past studies like \hyperref[sec:4]{[4]} and \hyperref[sec:5]{[5]} show that firms are willing to circumnavigate laws to collect data and encrypting data by design ensures that no third party can have access without being granted so explicitly.
+The aim of using a cryptographic hash like SHA-3 (Bertoni et al.,\ 2012) is to ensure a uniquely random reference to the data to avoid out-of-channel data leaks.\ Kratos suggests using an element of randomness like a salt to generate the hash in order to have the ability to revoke tokens by regenerating randomness.\ Kratos enforces that all schools encrypt their data before creating access tokens to mitigate the risk of loss/theft of data.\ Past research (Zetter,\ 2016) shows that firms are willing to circumnavigate laws to collect data and encrypting data by design ensures that no third party can have access without being granted so explicitly.
 \bigbreak
-Kratos suggests that encryption be done at the student level but this can also be at the school level depending on existing frameworks and rules surrounding the school.
-If the school chooses to encrypt student data on behalf of the student, Kratos enforces that the school use a unique key for each individual to minimize the risk of key theft.
-In addition to this, Kratos suggests that schools store their encrypted data in a distributed file storage system like IPFS to ensure data redundancy and availability in case of a setback.
-Storing data on IPFS also makes it easier to create timestamps since it is sufficient to reference the IPFS pointer instead of potentially hashing the whole data.
+Kratos suggests that encryption be done at the student level but this can also be at the school level depending on existing frameworks and rules surrounding the school.\ If the school chooses to encrypt student data on behalf of the student,\ Kratos enforces that the school use a unique key for each individual to minimize the risk of key theft.\ In addition to this,\ Kratos suggests that schools store their encrypted data in a distributed file storage system like IPFS to ensure data redundancy and availability in case of a setback.\ Storing data on IPFS also makes it easier to create time-stamps since it is sufficient to reference the IPFS pointer instead of potentially hashing the whole data.
 \bigbreak
-In the event a user wants to revoke access to a particular vendor, he could do so by changing the encryption key or by changing the randomness used to generate the access token.
-We suggest users do not regenerate encryption keys but Kratos will provide users an option to choose between the two.
+In the event that a user wants to revoke access to a particular vendor,\ he could do so by changing the encryption key or by changing the randomness used to generate the access token.\ We suggest that users do not regenerate encryption keys but Kratos will provide users an option to choose between the two.\ As we describe school data of K12 students in principle it is the school or school district administrator who controls such access.\ The difference is that the system will further establish accountability and auditability to students with the capacity to gain further control over their data when they become legally independent.
 \bigbreak
-Since schools have their own sets of policies, Kratos does not strictly enforce a set of practices for users to follow.
-This ensures that adoption of Kratos is not constrained by a certain set of rules.
-At the same time, Kratos defines a set of minimum requirements to be on board to ensure good practices on data protection are followed.
+Furthermore,\ since schools have their own sets of policies,\ Kratos does not strictly enforce a set of practices for users (at school or district level) to follow.\ This ensures that adoption of Kratos is not constrained by a certain set of rules.\ At the same time,\ Kratos defines a set of minimum requirements to be on board to ensure good practices on data protection are followed.
 \bigbreak
-Kratos also does not enforce how commitments need to be generated and stored, leaving it as an option for schools to provide their feedback on.
-Solutions provided by Kratos would include a centralized time-stamping server, a permissioned blockchain with the different schools as the nodes and simple time-stamping commitments to an existing blockchain.
-All three have their benefits and constraints and Kratos would enable schools to customize this to their requirements.
+Kratos also does not enforce how commitments need to be generated and stored,\ leaving it as an option for schools to provide their feedback.\ The system we propose would include a centralized time-stamping server,\ a permissioned blockchain with the different schools as the nodes,\ and simple time-stamping commitments to an existing blockchain.\ All three have their benefits and constraints and Kratos would enable schools to customize this to their requirements.
 \bigbreak
@@
 \subsection{Student agency and participation}
-The UNESCO framework for educational planning states that "the concern of planners is twofold: to reach a better understanding of the validity of education in its own empirically observed specific dimensions and to help in defining appropriate strategies for change" (Haddad, 1995, pp.
-5-6).
+The UNESCO framework for educational planning states that "the concern of planners is twofold: to reach a better understanding of the validity of education in its own empirically observed specific dimensions and to help in defining appropriate strategies for change" (Haddad,\ 1995,\ pp.\ 5-6).
 \bigbreak
-While summative and cumulative assessments provide "empirically observed specific dimensions" about student academic performance, student agency and active participation is equally required in order for policy and education to design "strategies for change".
-Growing use of education technologies enables data-driven decision-making [Gibson et al., 2015].
-Technology-mediated instruction and assessment tools with learning analytics functionality track and diagnose student progress.
-Most educational technologies can interpret and equip educators with information via digital dashboards and 'skill meters'[Baker, 2016; New, 2016].
-Fine-grained and continuously accumulated data about student behavior and performance surpasses traditional notions of assessment [Zeide, 2017] posing limitations over student dimensions of learning and equitable participation in the curriculum design.
+While summative and formative assessments provide "empirically observed specific dimensions" about student academic performance,\ student agency and active participation is equally required in order for policy and education to design "strategies for change".\ Growing use of education technologies enables data-driven decision-making (Gibson et al.,\ 2015).\ Technology-mediated instruction and assessment tools with learning analytics functionality track and diagnose student progress (Tempelaar et al.,\ 2013).\ Most educational technologies can interpret and equip educators with information via digital dashboards and 'skill meters'(Baker,\ 2016).\ Fine-grained and continuously accumulated data about student behavior and performance surpasses traditional notions of assessment (Zeide,\ 2017) posing limitations over student dimensions of learning and equitable participation in the curriculum design.
 \bigbreak
-Our prototype provides a graphical interface for student involvement in and accessibility to school data.
-While students and equally their parents can become acquainted with any changes and meanings of school data, students can also participate with personal feedback to the learning process.
-Student participation with personal perspectives and reflection are integral to the learning process (Ackermann, 1996).
-Some reflection is invariably carried out through questionnaire surveys examining school climate (Holahan and Batey, 2019).
-Both school climate surveys and the proposed student feedback application provide flexible selection of questionnaires and measurements with a common goal to improve school climate.
-However, the proposed application enables not only feedback from older students (Ibid.) but from all students, provided that the feedback addresses the goal to encourage perspective-taking and reflection that are deemed necessary to the learning process (Kegan, 1982).
-The application further enables student agency and control over the frequency, depth, and nature of the feedback provided that it directly reflects the learning experience.
+Our prototype provides a graphical interface for student involvement in (Figures 3-8) and accessibility to school data.\ While students and equally their parents can become acquainted with any changes and meanings of school data,\ students can also participate with personal feedback in the learning process (Figure 8).\ Student participation with personal perspectives and reflection are integral to the learning process (Ackermann,\ 1996).\ Some reflection is invariably carried out through questionnaire surveys examining school climate (Holahan and Batey,\ 2019).\ Both school climate surveys and the proposed student feedback application provide flexible selection of questionnaires and measurements with a common goal to improve school climate.\ However,\ the proposed application enables not only feedback from older students (Ibid.) but from all students,\ provided that the feedback addresses the goal to encourage perspective-taking and reflection that support learning (Kegan,\ 1982).\ The application further enables student agency and control over the frequency,\ depth,\ and nature of the feedback provided that it directly reflects the learning experience.
 
@@
 \subsection{Data literacy}
-Data literacy is the ability to understand, create, and communicate data as information.
-Beyond that, in increasingly data-driven systems it becomes imperative not only to understand data but to interact with it and participate in the decision-making processes that it increasingly begins to impact.
-While data provides insight about how systems operate, the inferences drawn from it is without context.
-Fine-grained data collected over long periods of time can lead to negative impact on individual well-being and pose limitations over future opportunities [Gasser et al., 2018].
-The growing digitization of schools create learning environments that enable constant data generation, the access to and use of which becomes hard to control, audit, and account for.
-The risk of digitized school environments constantly generating data and algorithmic assessments becoming permanent record that at any one point in time may limit individual opportunities leads us to prioritise on developing a socio-technological solution that enables applied data literacy for students at a young age.
-\textit{a note on Parents for Privacy Consortium and Data Quality Standards who make such efforts to educate students and parents about school data and that we (Kratos) complement these efforts.}
+ Data literacy is still a debatable construct (Bowler et al.,2017).\ Data means quantified information,\ which is "situated,\ taking its meaning from its context and the perspective of its beholder" (Borgman,\ 2015,\ p.\ 18).\ Making sense of data is relative.\ What data may mean to someone can mean nothing to others (Bowler et al.,,\ 2017).\ In increasingly data-driven systems it becomes imperative not only to understand data but to interact with it and participate in the decision-making processes that it increasingly begins to impact.\ While data provides insight about how systems operate,\ the inferences drawn from it can be without context.\ Fine-grained data collected over long periods of time can lead to negative impact on individual well-being and pose limitations over future opportunities (Altman et al.,\ 2018; Friedman and Nissenbaum,\ 1996).\
+ \bigbreak
+ The growing digitization of schools create learning environments that enable constant data generation,\ the access to,\ and use of which becomes hard to control,\ audit,\ account for,\ and understand its impact.\ The risk from digitized school environments that constantly generate data into a permanent record (Sirota,\ 2013) that at any one point in time may limit individual opportunities leads us to prioritise on developing a techno-social solution that enables applied data literacy for students at an ever younger age.
+ \bigbreak
+ While some studies examine aspects of data literacy in young people's lives (Bowler et al.,\ 2017; Selwyn and Pangrazio,\ 2018; Kumar et al.,\ 2017),\ the focus remains on data generated from the use of the Internet in general and social media specifically.\ Studies related to young people's perceptions,\ knowledge,\ and understanding of school data literacy are meagre.\
+ \bigbreak
+ Existing literature demonstrates that young people have varying interprations and a general understanding of data (Bowler et al.,\ 2017).\ However,\ many still find difficulties to connect with data at a concrete level,\ with the notion of having a data dossier (Ibid.,\ p.\ 27).\ This highlights the pressing need to develop strategies for school-related data literacy,\ introduce data management skills,\ and encourage student participation in an increasingly data-driven decision-making school environment.
+ \bigbreak
+ Kratos thus proposes a comprehensive,\ yet simple access,\ knowledge,\ and interaction with school data.\ At this conceptualisation stage,\ we choose three core data buckets to work with.\ These are assessment,\ including test scores,\ reports,\ and credentialing; conduct,\ including observation reports,\ behaviour,\ and attendance; and personally identifiable information,\ including personal identifiers,\ demographics,\ parent information,\ medical reports,\ and socio-economic information.\ We take these three data buckets to be crucial to student privacy and control.\
+ \bigbreak
+ In our conceptual work (Figures 3-5) we recognise that while schools collect data for each of these three data buckets,\ digital applications used in class also collect similar such data.\ For example,\ as schools contain data about conduct,\ applications such as Class Dojo (Manolev et al.,\ 2018) gather similar data.\ Teachers may further report behaviour using Swivl,\ a video application that enables supervision and classroom observation.\ Currently,\ students do not have a comprehensive access to or knowledge of all data that exists about their conduct.\ Moreover,\ there is no awareness and knowledge of the potential impact of all data that exists about their conduct.\ We therefore create visibility to the disparate data that exists about student conduct,\ provide visibility,\ and learning about such data (Figure 6).
 
@@
     \includegraphics[height=9.5cm]{{"images/data-detail-1.png}}
-    \caption{Data Information}
+    \caption{Data view}
     \label{fig:my_label}
@@
     \includegraphics[height=9.5cm]{{"images/data-detail-2.png}}
-    \caption{Detailed info about fields}
+    \caption{Data view with definition}
     \label{fig:my_label}
@@
 \section{Discussion and future work}
-In this paper we propose Kratos, a decentralised data management system which enables data interoperability, applied data literacy, and student participation in the curriculum design.
-This proposal envisions an ideal techno-social solution that empowers students by providing more control over their school data through data transparency, accountability and auditability.
-At the same time, continuous legal (GDPR, 2016; Colorado Student Data Transparency and Security Act, 2016; Kelly, 2019) and policy (Student Data Privacy Consortium, 2018) efforts are being made that will further help establish Kratos as a tool integral to the school ecosystem.
+In this paper we propose Kratos,\ a decentralised data management system,\ which enables data interoperability,\ applied data literacy,\ and student participation.\ This proposal envisions a techno-social solution that empowers students by providing control over their school data through data transparency and auditability.\ While legal (Colorado Student Data Transparency and Security Act,\ 2016; Kelly,\ 2019) and policy (e.g.\ SDPC,\ 2018) efforts are being made to ensure data privacy and security,\ much more needs to be done (Bulger et al.,\ 2019),\ which encourages us to continue work on the Kratos project as a technological layer of protection and data privacy.
 \bigbreak
-Unlike other sectors such as retail (Lu and Xu, 2017), finance (Ito et al., 2017), and medicine (Acbo et al., 2019), education is still not fully leveraging the power of global digital transformation.
-In medicine, for instance, data interoperability is important in order for various stakeholders to access comprehensive patient records and decentralised technologies have been used to enable such access while prioritising patient agency, privacy and security (Acbo et al., 2019).
+Unlike other sectors such as retail (Lu and Xu,\ 2017),\ finance (Ito et al.,\ 2017),\ and medicine (Agbo et al.,\ 2019),\ education is still not fully leveraging the power of global digital transformation.\ In medicine,\ for instance,\ data interoperability is important in order for various stakeholders to access comprehensive patient records.\ Across Europe (Demertzis et al.,\ 2018) and the United States (Azaria et al.,\ 2016),\ decentralised architectures have been developed to enable data sharing,\ interoperability,\ and access while prioritising patient agency,\ privacy,\ and security.\ Similarly,\ a more comprehensive data management for interoperability is highly needed in the education sector without compromising student privacy and control over their data.
 \bigbreak
-On the other hand, while data analytics provides insight about school processes, not everything that is measured is meaningful.
-Students’ perspectives and reflection should equally count next to data-driven decision-making.
-Moreover, the sophistication of data analytics and long-term data collection further decrease an individual's privacy.
-The risk of being publicly exposed can lead to being embarrassed or discriminated (Altman et al.
-2018).
-Constantly being aware of such risks can diminish one’s sense of freedom for self-expression and in the case with children and young people - the freedom to try out new things and make mistakes.
-Therefore, a decentralised data management system that prioritises student privacy and control over data, becomes crucial to the development of a safe and free learning environment.
+While our conceptual work focuses on K12 education we understand the complexities surrounding the sheer volume of data that is generated across many different systems,\ the differing interests of stakeholders in the education sector,\ and the hurdles involving access to data that many vendors consider proprietary.
 \bigbreak
-Future work includes three distinctive steps.
-First, we plan to develop a comprehensive taxonomy of school data.
-This will serve to develop a comprehensive data vocabulary for applied data literacy and identify data elements.
-The second step involves prototyping and formalising platform sections and functionality.
-And the third involves carrying out user studies to get student feedback and the fine tuning of the platform's user interface.
+We propose Kratos as a decentralised system that enables data integrity and visibility of the huge data trail of students who are still dependent on adults to decide for them.\ However,\ such a system can give them an immutable log of their data as an additional layer of security now and at any point in time beyond their secondary and higher education.
+\bigbreak
+Comprehensive access to data opens the possibilities for better analysis of processes and systems (Baker and Siemens,\ 2014).\ However,\ while data analytics gives some insight,\ not everything that is measured is valuable and not everything that is of value can be measured.\ Students’ perspectives should go on a par with data-driven decision-making.\ Moreover,\ the sophistication of data analytics and long-term data collection can potentially decrease individual privacy (Dwork and Rothblum,\ 2016).\ The risk of being publicly exposed can lead to embarrassment or discrimination (Altman et al.\ 2018).\ Constantly being aware of such risks can diminish one’s sense of freedom for self-expression and in the case with children and young people - the freedom to try out new things and make mistakes,\ things that are integral to learning.\ Therefore,\ a decentralised data management system that prioritises student privacy and control over data,\ becomes crucial to the development of a safe and free learning environment.
+\bigbreak
+Future work includes three distinctive steps.\ First,\ we plan to develop a comprehensive taxonomy of school data.\ This will serve to develop a comprehensive data vocabulary for applied data literacy and identify data elements - the trail and origin,\ and its purpose of use.\ The second step involves prototyping and formalising platform sections and functionality.\ And the third involves carrying out user studies to get student feedback and fine-tuning the platform's user interface.\ All future work is part of the greater effort to provide data privacy protection of students.\
 
 \section{Acknowledgements}
-We thank Cambridge Public School district, Student Data Privacy Consortium, Acess for Learning (SIF) for their support.
-We would also like to thank Rayner Ng Jing Kai and Goeffrey Martin from Yale-NUS College for their help in designing mockup designs and giving feedback on data interoperability.
+We thank Cambridge Public School district,\ Student Data Privacy Consortium,\ and Access for Learning (SIF) for their support,\ constant,\ and timely feedback.\ We would also like to thank Rayner Ng Jing Kai and Geoffrey Martin from Yale-NUS College for their help in sketching mock-up designs and giving feedback on data interoperability.\ Lastly we greatly appreciate the invaluable feedback on decentralised ledger technologies from Mario Galea at Random Consulting.
 
@@
     \label{sec:1}
-    \item Todd, P. OpenTimestamps: Scalable, Trustless, Distributed Timestamping with Bitcoin (2016). URL \url{https://petertodd.org/2016/opentimestamps-announcement}
+    \item Todd,\ P.\ OpenTimestamps: Scalable,\ Trustless,\ Distributed Timestamping with Bitcoin (2016).\ URL \url{https://petertodd.org/2016/opentimestamps-announcemen}
     \label{sec:2}
-    \item Szalachowski, P. (2018). Towards more reliable Bitcoin timestamps. arXiv preprint arXiv:1803.09028.
+    \item Szalachowski,\ P.\ (2018).\ Towards more reliable Bitcoin timestamps.\ arXiv preprint arXiv:1803.09028.
     \label{sec:3}
-    \item Massias, H., Avila, X. S., \& Quisquater, J. J. (1999). Design of a secure timestamping service with minimal trust requirement. In the 20th Symposium on Information Theory in the Benelux.
+    \item Massias,\ H.,\ Avila,\ X.\ S.,\ \& Quisquater,\ J.\ J.\ (1999).\ Design of a secure timestamping service with minimal trust requirement.\ In the 20th Symposium on Information Theory in the Benelux.
     \label{sec:4}
-    \item Zetter, K. Google Collected Data on Schoolchildren without permission (2016). URL \url{https://www.wired.com/2015/12/google-collected-data-on-schoolchildren-without-permission/}
+    \item Zetter,\ K.\ Google Collected Data on Schoolchildren without permission (2016).\ URL \url{https://www.wired.com/2015/12/google-collected-data-on-schoolchildren-without-permission/}
     \label{sec:5}
-    \item Dissent, Back-To-School Revolt in Springfield?
-Employees balk over using Google Drive as evidence of massive privacy breach mounts (2018).
-URL \url{https://www.pogowasright.org/back-to-school-revolt-in-springfield-employees-balk-over-using-google-drive-as-evidence-of-massive-privacy-breach-mounts/}
+    \item Singer,\ N.\ (2017).\ How Google took over the classroom.\ Available from: https://www.nytimes.com/2017/05/13/technology/google-education-chromebooks-schools.html
     \label{sec:6}
-    \item Bertoni, G., Daemen, J., Peeters, M., Van Assche, G., \& Van Keer, R. (2012). Keccak implementation overview. URL  \url{http://keccak.neokeon.org/Keccak-implementation-3.2.pdf}.
+    \item Bertoni,\ G.,\ Daemen,\ J.,\ Peeters,\ M.,\ Van Assche,\ G.,\ \& Van Keer,\ R.\ (2012).\ Keccak implementation overview.\ URL  \url{http://keccak.neokeon.org/Keccak-implementation-3.2.pdf}
     \label{sec:7}
-    \item CPSD. District Agreements Listing (2019). URL  \url{https://sdpc.a4l.org/district_listing.php?districtID=457}
+    \item CPSD.\ District Agreements Listing (2019).\ URL  \url{https://sdpc.a4l.org/district_listing.php?districtID=457}
     \label{sec:8}
-    \item Colins, A., and Halverson, R. (2018).\textit{ Re-thinking education in the age of technology: The digital revolution and schooling in America}. Teach Zeide, E. (2016).
+    \item Colins,\ A.,\ and Halverson,\ R.\ (2018).\textit{ Re-thinking education in the age of technology: The digital revolution and schooling in America}.\ Teachers College Press: New York.
     \label{sec:9}
-    \item 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?.
+    Zeide,\ E.\ (2017).\ 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?.
     \label{sec:10}
-    \item Colins, A., and Halverson, R. (2018).\textit{ Re-thinking education in the age of technology: The digital revolution and schooling in America}. Teachers College Press: New York.
+    \item Cody,\ A.\ (2013).\ Will the data warehouse become every student and teacher's 'permanent record'? \textit{Education Week},\ May,\ 20.
     \label{sec:11}
-    \item Cody, A. (2013). Will the data warehouse become every student and teacher's 'permanent record'? \textit{Education Week}, May, 20.
+    \item Gibson,\ D.\ C.,\ Webb,\ M.,\ and Ifenthaler,\ D.\ (2015).\ Challenges of big data in educational assessment.\textit{Proceedings of the IADIS International Conference of Exploratory Learning in Digital Age},\ 92-100.
     \label{sec:12}
-    \item Gibson, D.
-C., Webb, M., and Ifenthaler, D.
-(2015).
-Challenges of big data in educational assessment.\textit{Proceedings of the IADIS International Conference of Exploratory Learning in Digital Age}, 92-100.
+    \item Sultan,\ N.\ (2010).\ Cloud computing for education: a new dawn? 30 International Journal of Information Management  109.
     \label{sec:13}
-    \item Sultan, N. (2010). Cloud computing for education: a new dawn? 30 International Journal of Information Management  109.
+    \item Fitzgerald,\ B.\ (2014).\ Data collection isn't new.\ And it predates common core.Funny Monkey.
     \label{sec:14}
-    \item Fitzgerald, B. (2014). Data collection isn't new. And it predates common core.Funny Monkey.
+    \item Keegan,\ R.\ (1982).\ The evolving self: Problem and process in human development.\ Cambridge: Harvard UP.
     \label{sec:15}
-    \item Zeide, E. (2016). 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?
-    \item Kegan, R. (1982). \textit{The evolving self: Problem and process in human development}. Cambridge: Harvard University Press.
-    \item Ackermann, E.
-(1996).
-Perspective-Taking and object Construction.
-\textit{In Constuctionism in Practice: Designing, Thinking, and Learning in a Digital World} (Kafai, Y., and Resnick, M., Eds.).
-Mahwah, New Jersey: Lawrence Erlbaum Associates.
-Part 1, Chap.
-2.
-pp.
-25-37.
-    \item Gibson D.C., Webb M.E., Ifenthaler D.
-(2019) Measurement Challenges of Interactive Educational Assessment.
-In: Sampson D., Spector J., Ifenthaler D., Isaías P., Sergis S.
-(eds) Learning Technologies for Transforming Large-Scale Teaching, Learning, and Assessment.
-Springer, Cham.
-    \item Agbo, C. , Mahmoud, Q., and Eklund, J. (2019). Blockchain technology in healthcare: A systematic review. \textit{Healthcare, 7\textit{ (56)}}, 1-30.
-    \item Lu, Q. and Xu, X. (2017). Adaptable blockchain-based systems: A case study for product traceability. \textit{IEEE Software, 34\textit{ (6)}}, 21-27.
-    \item Ito, J., Narula, N., and Ali, R.
-(2017).
-The blockchain will do to the financial system what the Internet did to the media.
-\textit{Harvard Business Review}, March 23.
-Available from: https://hbr.org/2017/03/the-blockchain-will-do-to-banks-and-law-firms-what-the-internet-did-to-media
-    \item U.S.
-Department of Education, Office of Educational Technology (2012).
-\textit{Enhancing teaching and learning through educational data mining and learning analytics: An issue brief}.
-U.S.
-Department of education: Washington, D.C.
-    \item EU General Data Protection Regulation (GDPR): Regulation (EU) 2016/679 of the European Parliament and of the Council of 27 April 2016 on the protection of natural persons with regard to the processing of personal data and on the free movement of such data, and repealing Directive 95/46/EC (General Data Protection Regulation), OJ 2016 L 119/1.
-    \item Kelly, M. (2019). New privacy bill would give parents an ‘Eraser Button’ and ban ads targeting children. \textit{The Verge}, March 12.
-
+    \item Ackermann,\ E.\ (1996).\ \textit{Constructionism in practice: designing,\ thinking,\ and learning in a digital world}.\ Routledge.
+    \label{sec:16}
+    \item Gibson,\ D.\ C.,\ Webb,\ M.\ E.,\ and Ifenthaler,\ D.\ (2019).\ Measurement Challenges of Interactive Educational Assessment.\ In \textit{Learning Technologies for Transforming Large-Scale Teaching,\ Learning,\ and Assessment},\ 19-33.\ Springer,\ Cham.
+    \label{sec:17}
+    \item Agbo,\ C.\ C.,\ Mahmoud,\ Q.\ H.,\ and Eklund,\ J.\ M.\ (2019,\ June).\ Blockchain Technology in Healthcare: A Systematic Review.\ In \textit{Healthcare} (Vol.\ 7,\ No.\ 2,\ p.\ 56).\ Multidisciplinary Digital Publishing Institute.
+    \label{sec:18}
+    \item Lu,\ Q.,\ and Xu,\ X.\ (2017).\ Adaptable blockchain-based systems: a case study for product traceability.\ \textit{IEEE Software,} 34(6),\ 21-27.
+    \label{sec:19}
+    \item Ito,\ J.,\ Narula,\ N.,\ and Ali,\ R.\ (2017).\ The blockchain will do to the financial system what the internet did to media.\ \textit{Harvard Business Review,} 9(March).
+    \label{sec:20}
+    \item Bienkowski,\ M.,\ Feng,\ M.,\ and Means,\ B.\ (2012).\ Enhancing teaching and learning through educational data mining and learning analytics: An issue brief.\ \textit{US Department of Education,\ Office of Educational Technology,} 1,\ 1-57.\
+    \label{sec:21}
+    \item Regulation,\ G.\ D.\ P.\ R.\ (2016).\ Regulation (EU) 2016/679 of the European Parliament and of the Council of 27 April 2016 on the protection of natural persons with regard to the processing of personal data and on the free movement of such data,\ and repealing Directive 95/46.\ \textit{Official Journal of the European Union (OJ)},\ 59(1-88),\ 294.
+    \label{sec:22}
+    \item Kelly,\ M.\ (2019).\ New privacy bill would give parents an ‘Eraser Button’ and ban ads targeting children.\ URL \url{https://www.theverge.com/2019/3/12/18261181/eraser-button-bill-children-privacy-coppa-hawley-markey}
+    \label{sec:23}
+    \item Demertzis,\ I.,\ Papadopoulos,\ S.,\ Papapetrou,\ O.,\ Deligiannakis,\ A.,\ Garofalakis,\ M.,\ and Papamanthou,\ C.\ (2018).\ Practical Private Range Search in Depth.\ \textit{ACM Transactions on Database Systems,\ 43(1),} 2.
+    \label{sec:24}
+    \item Altman,\ M.,\ Wood,\ A.\ B.,\ O'Brien,\ D.,\ and Gasser,\ U.\ (2018).\ Practical approaches to big data privacy over time.
+    \label{sec:25}
+    \item Bowler,\ L.,\ Acker,\ A.,\ Jeng,\ W.,\ and Chi,\ Y.\ (2017).\ It lives all around us: Aspects of data literacy in teen’s lives.\ \textit{Proceedings of the Association for Information Science and Technology,\ 54(1)},\ 27–35.\ DOI:10.1002/pra2.2017.14505401004
+    \label{sec:26}
+    \item Friedman,\ B.,\ and Nissenbaum,\ H.\ (1996).\ Bias in computer systems.\ \textit{ACM Transactions on Information Systems (TOIS),\ 14(3)},\ 330-347.
+    \label{sec:27}
+    \item Borgman,\ C.\ L.\ (2015).\ \textit{Big data,\ little data,\ no data: Scholarship in the networked world.} MIT Press.
+    \label{sec:28}
+    \item Kumar,\ P.,\ Naik,\ S.,\ Devkar,\ U.,\ Chetty,\ M.,\ Tamara,\ C.,\ and Vitak,\ J.\ (2017).\ 'No telling passcodes out because they’re private’: Understanding children’s mental models of privacy and security online.\ \textit{Proceedings of the ACM on Human-Computer Interaction,\ 1(CSCW)},\ 64.
+    \label{sec:29}
+    \item Azaria,\ A.,\ Ekblaw,\ A.,\ Vieira,\ T.,\ and Lippman,\ A.\ (2016,\ August).\ MedRec: Using blockchain for medical data access and permission management.\ \textit{In 2016 2nd International Conference on Open and Big Data (OBD)},\  25-30.\ IEEE.
+    \label{sec:30}
+    \item Zeide,\ E.\ (2014).\ The proverbial permanent record.
+    \label{sec:31}
+    \item Common Education Data Standards - CEDS 101.\ Available from: https://ceds.ed.gov/pdf/ceds-101.pdf
+    \label{sec:32}
+    \item Tempelaar,\ D.\ T.,\ Heck,\ A.\ J.\ P.,\ Cuypers,\ H.,\ van der Kooij,\ H.,\ and van de Vrie,\ E.\ (2013).\ Formative assessment and learning analytics.\ in D.\ Suthers,\ K.\ Verbert,\ E.\ Duval,\ and X.\ Ochoa (Eds.),\ LAK 2013: Third International Conference on Learning Analytics and Knowledge: 8-12 April,\ 2013,\ Leuven,\ Belgium,\ 205-209.\ New York: Association for Computing Machinery.\
+    \label{sec:33}
+    \item Bulger,\ M.,\ Collins,\ S.,\ Sallay,\ D.,\ Vance,\ A.\ (2019).\ School safety and privacy: An animated introduction.\ \textit{FERPA,\ SHERPA: The Education Privacy Resource Centre}.\ Available from: https://ferpasherpa.org/schoolsafetyvideo/
+    \label{34}
+    \item Baker,\ R.\ and Siemens,\ G.\ (2014).\ Educational data mining and learning analytics.\ In Sawyer K.\ (Ed.),\ Cambridge handbook of the learning sciences (2nd ed.),\ 253–274.\ New York,\ NY: Cambridge University Press.
+    \label{35}
+    \item Snyder,\ T.D.,\ de Brey,\ C.,\ and Dillow,\ S.A.\ (2016).\ Digest of Education Statistics 2015 (NCES 2016-014).\ National Center for Education Statistics,\ Institute of Education Sciences,\ U.S.\ Department of Education: Washington,\ DC.
+    \label{36}
+    \item Militello,\ M.,\ Bass,\ L.,\ Jackson,\ T.\ K.,\ and Wang,\ Y.\ (2013).\ How data are used and misused in schools: Perceptions from teachers and principals.\textit{Education Sciences,\ 3},\ 98-120.
+    \label{37}
+    \item Arnold,\ K.\ E.\ and Pistilli,\ M.\ D.\ (2012,\ April).\ Course signals at Purdue: Using learning analytics to increase student success.\ Proceedings of the 2nd International Conference on Learning Analytics and Knowledge,\ 267–270.
+    \label{38}
+    \item Baker,\ R.\ S.\ (2016).\ Stupid tutoring systems,\ intelligent humans.\ \textit{International Journal of Artificial Intelligence in Education,\ 26},\ 600-614.
+    \label{39}
+    \item Kelly,\ G.,\ Graham,\ J.,\ and Fitzgerald,\ B.\ (2018).\ \textit{State of edtech privacy report.} Common Sense Privacy Evaluation Initiative.\ San Francisco,\ CA: Common Sense.\ 
+    \label{40}
+    \item Data Quality Campaign.\ (2018).\ What parents and teachers think about education data.\ Available from: https://dataqualitycampaign.org/resource/what-parents-and-teachers-think-about-education-data/
+    \label{41}
+    \item Student Data Privacy Consortium.\ (2018).\ Student Data Privacy Consortium: Policy and procedures.\ SDPC.\ Available from: https://privacy.a4l.org/wp-content/uploads/2018/06/Student-Data-Privacy-Consortium-P-and-P-2018-Final.pdf
+    \label{42}
+    \item California Consumer Privacy Act of 2018 (CCPA),\ CAL.\ CIV.\ CODE § 1798.198(a) (2018).
+    \label{43}
+    \item Colorado Student Data Transparency and Security Act (HB 16-1423).\ (2016).\ Available from: https://www.cde.state.co.us/dataprivacyandsecurity/crs22-16-101
+    \label{44}
+    \item Agbo,\ C.,\ Mahmoud,\ Q.,\ and Eklund,\ J.\ (2019).\ Blockchain technology in healthcare: A systematic review.\ \textit{Healthcare,\ 7},\ 56,\ 1-30.
+    \label{45}
+    \item Dwork,\ C.,\ and Rothblum,\ G.\ (2016).\ Concentrated differential privacy.\ Working paper.\ Available from: https://arxiv.org/pdf/1603.01887.pdf
+    \label{46}
+    \item Sirota,\ D.\ (2013).\ Big data means kids' "permanent records" might never be erased.\ MOTHERBOARD,\ October 24.\ Available from: http://motherboard.vice.com/blog/permanent-records-are-hurting-kids.
+    \label{47}
+    \item Selwyn,\ N.,\ and Pangrazio,\ L.\ (2018).\ Doing data differently?
+Developing personal data tactics and strategies amongst young mobile media users.\ \textit{ Big Data and Society},\ January-June,\ 1-12.
+    \label{48}
+    \item Manolev,\ J.\ Sullivan A.,\ and Slee,\ R.\ (2018).\ The datafication of discipline: Class Dojo,\ surveillance and a performative classroom culture.\ Learning,\ Media and Technology,\ 44(1),\ 36-51.
 \end{enumerate}
-
 \end{document}

### 2019-06-07 paper.tex
add submission draft

@@
 \textit{Abstract}
-Growing digitization has made data ownership an important focus point for institutions and students.\ Broadly,\ there are three issues,\ which require urgent attention for optimization of data privacy,\ literacy,\ and utilization.\ First,\ schools globally are equivocal about data generated by and about students as a result of the digitization of instruction,\ learning,\ and assessment.\ They lack necessary frameworks for data literacy,\ data interoperability,\ and optimization,\ while maintaining privacy and control.\ Second,\ the scale,\ source,\ and nature of school data makes its interoperability impractical,\ resulting in an inability to assess the true impact of educational technologies on instruction and learning.\ Third,\ while data helps teachers improve pedagogical practices,\ an increasingly data-driven decision-making process suggests that student dimensions of learning and equitable participation in curriculum design becomes secondary.\ Finding a balance between data-driven decision making and student voice is critical for an efficient school ecosystem.\ In this paper,\ we introduce Kratos: an immutable decentralised data management system that provides data privacy and applied data literacy while empowering students with a user interface for data governance and active participation in the school ecosystem.\ Using the advantages of blockchain technologies,\ Kratos enables easy authentication and access to data.\ The objective of Kratos is thus to equip schools and students with the ability to access,\ manage and control their data and to understand how,\ why,\ and by whom data is accessed without compromising student agency and privacy.\ This paper describes proof of concept for Kratos and its benefits to the school ecosystem.
+Growing digitization has made data ownership an important focus point for institutions and students.\ Broadly,\ there are three issues,\ which require urgent attention for optimization of data privacy,\ literacy,\ and utilization.\ First,\ schools globally are equivocal about data generated by and about students as a result of the digitization of instruction,\ learning,\ and assessment.\ They lack necessary frameworks for data literacy,\ data interoperability,\ and optimization,\ while maintaining privacy and control.\ Second,\ the scale,\ source,\ and nature of school data makes its interoperability impractical,\ resulting in an inability to assess the true impact of educational technologies on instruction and learning.\ Third,\ while data helps teachers improve pedagogical practices,\ an increasingly data-driven decision-making process suggests that student dimensions of learning and equitable participation in curriculum design becomes secondary.\ Finding a balance between data-driven decision making and student voice is critical for an efficient school ecosystem.\ In this paper,\ we introduce Kratos: an immutable decentralised data management system that provides data privacy and applied data literacy while empowering students with a user interface for data governance and active participation in the school ecosystem.\ Using the advantages of blockchain technologies,\ Kratos enables easy authentication and access to data.\ The objective of Kratos is thus to equip students and schools with the ability to access,\ manage, and control their data and to understand how,\ why,\ and by whom data is accessed without compromising student agency and privacy.\ This paper describes proof of concept for Kratos, its benefits to students and schools, and necessary future work.
 \bigbreak
@@
 
-To improve work,\ school practitioners need data interoperability and educational information from data mining (Bienkowski,\ Feng,\ and Means,\ 2012; Snyder et al.,\ 2016).\ This need emanates from the requirement to measure school effectiveness by combining student achievement,\ data,\ and accountability (Militello et al.,\ 2013).
+To improve work,\ school practitioners need data interoperability and educational information from data mining (Bienkowski,\ Feng,\ and Means,\ 2012; Snyder et al.,\ 2016).\ Broadly, this need emanates from the requirement to measure school effectiveness by combining student achievement,\ data,\ and accountability (Militello et al.,\ 2013).
 \bigbreak
@@
 \bigbreak
-Kratos suggests that encryption be done at the student level but this can also be at the school level depending on existing frameworks and rules surrounding the school.\ If the school chooses to encrypt student data on behalf of the student,\ Kratos enforces that the school use a unique key for each individual to minimize the risk of key theft.\ In addition to this,\ Kratos suggests that schools store their encrypted data in a distributed file storage system like IPFS to ensure data redundancy and availability in case of a setback.\ Storing data on IPFS also makes it easier to create time-stamps since it is sufficient to reference the IPFS pointer instead of potentially hashing the whole data.
+Kratos suggests that encryption be done at the student level but this can also be at the school level depending on existing frameworks and rules surrounding the school.\ If the school chooses to encrypt student data on behalf of the student,\ Kratos enforces that the school use a unique key for each individual to minimize the risk of key theft.\ In addition to this,\ Kratos suggests that schools store their encrypted data in a distributed file storage system like IPFS to ensure data redundancy and availability in case of a setback.\ Storing data on IPFS also makes it easier to create time-stamps since it is sufficient to reference the IPFS pointer instead of potentially hashing the whole data.\ It is to be noted that schools would not be burdened with the load of encrypting data and choosing between different levels of encryption would be a simple click on Kratos' interface.
 \bigbreak
